\input cwebmac
\datethis

\N{1}{1}Introduction. This program finds a minimum-move solution to the famous
``15 puzzle,'' using a method introduced by Richard E. Korf
[{\sl Artificial Intelligence\/ \bf27} (1985), 97--109]. It's the
first of a series of ever-more-efficient ways to do the job.
(You might want to read the zeroth program in the series,
{\mc 15PUZZLE-KORF0}, as background, although much of the
documentation is repeated here.)
My main reason for writing this group of routines was to experiment
with a new (for me) style of programming, explained below.

The initial position is specified on the command line as a permutation of the
hexadecimal digits
$\{\.0,\.1,\.2,\allowbreak\.3,\allowbreak\.4,\.5,\.6,\.7,\.8,\.9,\allowbreak
\.a,\.b,\.c,\.d,\.e,\.f\}$;
this permutation is used to fill the rows of a $4\times4$ matrix, from top to
bottom and left to right. For example, `\.{159d26ae37bf48c0}' specifies the
starting position
$$\vcenter{\halign{&\enspace\tt#\cr
1&5&9&d\cr
2&6&a&e\cr
3&7&b&f\cr
4&8&c&0\cr
}}$$
The number \.0 stands for a blank cell. Each step in solving the puzzle
consists of swapping \.0 with one of its neighbors. The goal position
is always
$$\vcenter{\halign{&\enspace\tt#\cr
1&2&3&4\cr
5&6&7&8\cr
9&a&b&c\cr
d&e&f&0\cr
}}$$
(Korf had a different goal position, namely \.{0123456789abcdef}. I agree
that his convention is mathematically superior to mine; but it conflicts
with a 125-year-old tradition. So I have retained the historic practice.
One can of course interchange our conventions, if desired, by
rotating the board by $180^\circ$ and by replacing each nonzero digit
$x$ by $16-x$.)

\Y\B\8\#\&{include} \.{<stdio.h>}\6
\8\#\&{include} \.{<time.h>}\6
\&{char} \\{board}[\T{16}];\6
\&{char} \\{start}[\T{16}];\6
\&{int} \\{stack}[\T{100}];\6
\&{int} \\{timer};\7
\\{main}(\&{int} \\{argc}${},\39{}$\&{char} ${}{*}\\{argv}[\,]){}$\1\1\2\2\6
${}\{{}$\1\6
\&{register} \&{int} \|j${},{}$ \|k${},{}$ \|s${},{}$ \|t${},{}$ \\{del}${},{}$
\\{piece}${},{}$ \\{moves};\7
\X2:Input the initial position\X;\6
\X5:Apply Korf's procedure\X;\6
\X24:Output the results\X;\6
\4${}\}{}$\2\par
\fi

\M{2}Let's regard the 16 cell positions as two-digit numbers in quaternary
(radix-4) notation:
$$\vcenter{\halign{&\enspace#\cr
00&01&02&03\cr
10&11&12&13\cr
20&21&22&23\cr
30&31&32&33\cr
}}$$
Thus each cell is identified by its row number $r$ and its column number $c$,
making a two-nyp code $(r,c)$, with $0\le r,c<4$.

Furthermore, it's convenient to renumber the input digits \.1, \.2,
\dots,~\.f, so that they match their final destination positions,
00, 01, \dots,~32. This conversion simply subtracts~1; so \.0 gets
mapped into $-1$. The example initial position given earlier
will therefore appear as follows in the \PB{\\{start}} array:
$$\vcenter{\halign{&\enspace$\hfil#\hfil$\cr
00&10&20&30\cr
01&11&21&31\cr
02&12&22&32\cr
03&13&23&-1\cr
}}$$

Half of the initial positions make the puzzle unsolvable, because the
permutation must be odd if and only if the \.0 must move an odd number of
times. This solvability condition is checked when we read the input.

\Y\B\4\D$\\{row}(\|x)$ \5
$((\|x)\GG\T{2}{}$)\par
\B\4\D$\\{col}(\|x)$ \5
$((\|x)\AND\T{\^3}{}$)\par
\Y\B\4\X2:Input the initial position\X${}\E{}$\6
\&{if} ${}(\\{argc}\I\T{2}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Usage:\ \%s\ startposi}\)\.{tion\\n"},\39%
\\{argv}[\T{0}]){}$;\5
${}\\{exit}({-}\T{1});{}$\6
\4${}\}{}$\2\6
\&{for} ${}(\|j\K\T{0};{}$ ${}\|k\K\\{argv}[\T{1}][\|j];{}$ ${}\|j\PP){}$\5
${}\{{}$\1\6
\&{if} ${}(\|k\G\.{'0'}\W\|k\Z\.{'9'}){}$\1\5
${}\|k\MRL{-{\K}}\.{'0'};{}$\2\6
\&{else} \&{if} ${}(\|k\G\.{'a'}\W\|k\Z\.{'f'}){}$\1\5
${}\|k\MRL{-{\K}}\.{'a'}-\T{10};{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"The\ start\ position\ }\)\.{should\ use\ only\
hex\ }\)\.{digits\ (0123456789ab}\)\.{cdef)!\\n"});{}$\6
${}\\{exit}({-}\T{2});{}$\6
\4${}\}{}$\2\6
\&{if} (\\{start}[\|k])\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Your\ start\ position}\)\.{\ uses\ \%x\ twice!%
\\n"},\39\|k){}$;\5
${}\\{exit}({-}\T{3});{}$\6
\4${}\}{}$\2\6
${}\\{start}[\|k]\K\T{1};{}$\6
\4${}\}{}$\2\6
\&{for} ${}(\|k\K\T{0};{}$ ${}\|k<\T{16};{}$ ${}\|k\PP){}$\1\6
\&{if} ${}(\\{start}[\|k]\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Your\ start\ position}\)\.{\ doesn't\ use\ %
\%x!\\n"},\39\|k){}$;\5
${}\\{exit}({-}\T{4});{}$\6
\4${}\}{}$\2\2\6
\&{for} ${}(\\{del}\K\|j\K\T{0};{}$ ${}\|k\K\\{argv}[\T{1}][\|j];{}$ ${}\|j%
\PP){}$\5
${}\{{}$\1\6
\&{if} ${}(\|k\G\.{'0'}\W\|k\Z\.{'9'}){}$\1\5
${}\|k\MRL{-{\K}}\.{'0'}{}$;\5
\2\&{else}\1\5
${}\|k\MRL{-{\K}}\.{'a'}-\T{10};{}$\2\6
${}\\{start}[\|j]\K\|k-\T{1};{}$\6
\&{for} ${}(\|s\K\T{0};{}$ ${}\|s<\|j;{}$ ${}\|s\PP){}$\1\6
\&{if} ${}(\\{start}[\|s]>\\{start}[\|j]){}$\1\5
${}\\{del}\PP{}$;\C{ count inversions }\2\2\6
\&{if} ${}(\|k\E\T{0}){}$\1\5
${}\|t\K\|j;{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(((\\{row}(\|t)+\\{col}(\|t)+\\{del})\AND\T{\^1})\E\T{0}){}$\5
${}\{{}$\1\6
\\{printf}(\.{"Sorry\ ...\ the\ goal\ }\)\.{is\ unreachable\ from\ }\)\.{that\
start\ position!}\)\.{\\n"});\6
\\{exit}(\T{0});\6
\4${}\}{}$\2\par
\U1.\fi

\N{1}{3}Korf's method.
If piece $(r,c)$ is currently in board position $(r',c')$, it must make
at least $\vert r-r'\vert+\vert c-c'\vert$ moves before it reaches home.
The sum of these numbers, over all fifteen pieces, is called~$h$;
this quantity is also known as the ``Manhattan metric lower bound.''

We will say that a move is {\it happy\/} if it moves a piece closer
to the goal; otherwise we'll call the move {\it sad}.
Korf's key idea is to try first to find a solution that makes only happy
moves. (For example, one can actually win from the starting position
\.{bc9e80df3412a756} by making $h=56$ moves that are entirely happy.)
If that fails, we start over, but this time we try to make $h+1$ happy moves
and 1~sad move. And if that also fails, we try for $h+k$ happy moves
and $k$ sad ones, for $k=2$, 3, \dots, until finally we succeed.

That strategy may sound silly, because each new value of~$k$ repeats
calculations already made. But it's actually brilliant, for two reasons:
(1)~The search can be carried out with almost no memory, for any
fixed value of~$k$ --- in fact, this program uses fewer than 500
bytes for all its data. (2)~The total running time for $k=0$, 1, \dots,~$k_0$
isn't much more than the time needed for a {\it single\/} run with
$k=k_0$, because the running time increases exponentially with~$k$.

Memory requirements are minimal because we can get away with memoryless
depth-first search to explore all solutions. The fifteen puzzle is
much nicer in this respect than many other problems; indeed, a blind
depth-first search as used here is often a poor choice in other applications,
because it might explore many subproblems repeatedly, having no inkling
that it has already ``been there and done that.'' But the search tree
of the fifteen puzzle has comparatively few overlapping branches.
For example, the shortest cycle of moves that restores a previously examined
position occurs only when we go thrice around a $2\times2$ subsquare,
leading to a cycle of length 12; therefore the first five levels below any
node of the tree consist entirely of distinct states of the board,
and only a few duplicates occur at level six.

Note: The Manhattan metric is somewhat weak, and substantially better
lower bounds are known. Some day I plan to incorporate them into
later programs in this series. The simple approach adopted here
is sufficient to handle most random initial positions in a reasonable
amount of time. But when it is applied to the example of transposition,
in the introduction, it is too slow: That example has a Manhattan
lower bound of 40, yet the shortest solutions have 72 moves. Thus
the $k$ value for that problem is 16; and each new value of~$k$
takes empirically about 5.7 times as long as the previous case.
(So the running time was 40.4 hours on my Opteron computer, vintage 2004.)

Moreover, I could improve the present scheme in various other ways.
Even with the Manhattan metric, I find that there are 88,728,779 positions
from which the goal is achievable without any sad moves. (A surprisingly
large number --- at least, I was expecting less than a million. Incidentally,
exactly 114 of those positions have the maximum distance to the goal,
namely 56, as in the example stated above.) If we had a
table of those positions in memory, we would know that our last-permitted
sad move must be to something in the table; that might make a
significant speedup. Korf and Taylor have described a series of
other improvements that actually
allowed them to solve random instances of the 24-puzzle in 1996
[{\sl AAAI National Conference Proceedings\/} (1996), 1202--1207].

Another note: Maybe my terminology ``happy'' versus ``sad'' is too
poignant. Should I have called them ``downhill'' and ``uphill'' moves
instead? Readers comments are invited.

\fi

\M{4}Saving memory means that the computation lives entirely within the
computer's high-speed cache. Of course we do need to examine billions
of positions, in tough cases; so we want the inner loop of depth-first
search to be short.

Let's take a look at the operations that are
involved when, say, we've just moved the empty cell into position
$(r,c)$, coming from the south. (In other words, we will
imagine that the empty cell was
previously at $(r+1,c)$. We'll think of the {\it empty\/} cell as moving north,
although the board has actually changed by moving a {\it piece\/} to the
south.)
From $(r,c)$ we'll try to move the empty cell either west to $(r,c-1)$,
or north to $(r-1,c)$, or east to $(r,c+1)$;
up to three moves are possible, depending on whether $(r,c)$ is a middle cell,
an edge cell, or a corner cell. After having tried all those possibilities,
we'll backtrack and make the next move from the cell to the south
that got us started.

A straightforward implementation appears in the program {\mc 15PUZZLE-KORF0},
where the inner loop contains many tests:
(1)~Choosing a direction. (2)~Testing if this direction is feasible
from the current $(r,c)$. (3)~If feasible, is the move happy or sad?
(4)~If it's sad, have we exceeded our quota of sad moves?
(5)~Have we reached the goal? (6)~Have we exhausted all possibilities?

Such conditional branches play havoc with the pipeline organization of
a modern computer. Therefore the present program avoids most of them
by straight-line coding, using essentially a finite-state automaton
to control the actions. The program consists conceptually of $4\times4\times
4\times4=256$ small parts, one for each combination of $(r,c,d,p)$,
where we're at empty cell position $(r,c)$ trying to move the empty
cell in direction~$d$, having entered this cell from direction~$p$.

(In fact, 256 is a generous upper estimate, because many combinations
are impossible at the edges and corners. We won't, for example, try
to move east when the empty cell is in column~3. The true number of
cases is 152, of which 48 simply undo a move that was made.)

I could have generated the individual pieces of code with the macro
capability of \CEE/'s preprocessor. But my debugging tools don't know
how to deal with macros satisfactorily. So I did the repetitive steps
with macros at another level, namely with \.{emacs} as I typed this
program into the computer. The \CEE/ preprocessor did, however, come in handy
to make the code more readable than it would have been if I had
fully expanded everything by text editing.

Instead of writing this program as a collection of 152 little modules,
I could have made it a collection of 152 little procedures. But that
wouldn't have bought me anything; as a full professor with tenure,
I don't have to worry about being fired when I use \PB{\&{goto}} statements.
The overhead of subroutine calling, and the fact that ``tail recursion''
is extremely useful in the situations met here, both convince me that
no state-of-the-art compiler would come up with anything near as efficient
if I wrote procedures instead of this code that purposely looks like
spaghetti. Nor would a procedure-oriented version be easier for me to write.

The only drawback to the style adopted here, as far as I can
see, is that I had many chances to make mistakes; a program of this kind
will seem to work even though it contains typographic errors.
Therefore I had to do a lot of careful desk checking.

Does all this replication help? The program now has shifted the
computer's work from the instruction cache to the data cache.
One natural benchmark is the case \.{ca6098dfb73254e1}, which
proved to be the toughest of the 100 random examples in Korf's
original paper. (He reported that his implementation
searched more than 6 billion nodes in that case.)
In my first tests, on an Athlon computer purchased in the year 2000,
the running time for this toughie was 8 minutes and 58 seconds.
By comparison, {\mc 15PUZZLE-KORF0} requires 24 minutes and 26 seconds
on the same machine. Korf's original experiment, using a Pascal
compiler on a {\mc DEC} 2060 in 1984, took about 4000 minutes;
so the present program represents a more than 400-fold speedup,
of which we can attribute a factor of roughly 160 to improvements
in hardware, and another factor of about 2.7 to the implementation
technique adopted here.

\fi

\M{5}Enough of this introduction. Let's get on with the program.

\Y\B\4\X5:Apply Korf's procedure\X${}\E{}$\6
\X6:Set \PB{\\{moves}} to the minimum number of happy moves\X;\6
\&{if} ${}(\\{moves}\E\T{0}){}$\1\5
\&{goto} \\{win};\C{ otherwise our solution will take $6+6$ moves! }\2\6
\&{while} (\T{1})\5
${}\{{}$\1\6
${}\\{timer}\K\\{time}(\T{0});{}$\6
${}\|t\K\\{moves}{}$;\C{ desired number of \PB{$(\hbox{(sad moves)}\LL\T{8})+%
\hbox{(happy moves)}$} }\6
\X23:Try for a solution with \PB{\|t} more moves\X;\6
${}\\{printf}(\.{"\ ...\ no\ solution\ wi}\)\.{th\ \%d+\%d\ moves\ (\%d\ s}\)%
\.{ec)\\n"},\39\\{moves}\AND\T{\^ff},\39\\{moves}\GG\T{8},\39\\{time}(\T{0})-%
\\{timer});{}$\6
${}\\{moves}\MRL{+{\K}}\T{\^101}{}$;\C{ add a sad move and a happy move to the
current quota }\6
\4${}\}{}$\2\6
\4\\{win}:\par
\U1.\fi

\M{6}\B\X6:Set \PB{\\{moves}} to the minimum number of happy moves\X${}\E{}$\6
\&{for} ${}(\|j\K\\{moves}\K\T{0};{}$ ${}\|j<\T{16};{}$ ${}\|j\PP){}$\1\6
\&{if} ${}(\\{start}[\|j]\G\T{0}){}$\5
${}\{{}$\1\6
${}\\{del}\K\\{row}(\\{start}[\|j])-\\{row}(\|j);{}$\6
${}\\{moves}\MRL{+{\K}}(\\{del}<\T{0}\?{-}\\{del}:\\{del});{}$\6
${}\\{del}\K\\{col}(\\{start}[\|j])-\\{col}(\|j);{}$\6
${}\\{moves}\MRL{+{\K}}(\\{del}<\T{0}\?{-}\\{del}:\\{del});{}$\6
\4${}\}{}$\2\2\par
\U5.\fi

\M{7}The main control routine is a stack, which records two things:
In the left 16 bits is the number of $\rm(sad,happy)$ moves remaining,
called~$t$; this number will be restored when backtracking.
And in the right 16 bits is a code number of the routine to
execute next, after finishing every task that's higher on the stack.

\fi

\M{8}\B\X8:Switch into action\X${}\E{}$\6
\4\\{switcher}:\5
${}\|t\K\\{stack}[\MM\|s]\GG\T{16};{}$\6
\&{switch} ${}(\\{stack}[\|s]\AND\T{\^ffff}){}$\5
${}\{{}$\1\6
\X10:Cases that move east from column 0\X\6
\X11:Cases that move east from column 1\X\6
\X12:Cases that move east from column 2\X\6
\X13:Cases that move west from column 1\X\6
\X14:Cases that move west from column 2\X\6
\X15:Cases that move west from column 3\X\6
\X16:Cases that move north from row 1\X\6
\X17:Cases that move north from row 2\X\6
\X18:Cases that move north from row 3\X\6
\X19:Cases that move south from row 0\X\6
\X20:Cases that move south from row 1\X\6
\X21:Cases that move south from row 2\X\6
\X22:Cases that move back\X\6
\4\&{case} \\{bottom}:\5
\&{break};\6
\4\&{default}:\5
${}\\{fprintf}(\\{stderr},\39\.{"Oops,\ I'm\ confused\ }\)\.{about\ case\ \%x!%
\\n"},\39\\{stack}[\|s]);{}$\6
\4${}\}{}$\2\par
\U23.\fi

\M{9}Directions are encoded as follows: $\rm east=0$, $\rm north=1$,
$\rm west=2$, $\rm south=3$. (Think of powers of~$i$ in the complex
plane.)

Each feasible combination of $(r,c,d,p)$ is encoded as an 8-bit quantity
called \PB{$\\{code}(\|r,\|c,\|d,\|p)$}, so that we can easily switch to it.
There's also \PB{$\\{tailcode}(\|r,\|c,\|d)$}; this stands for a case that will
try
only a single direction \PB{\|d} at cell $(r,c)$ before returning.

To avoid complicated testing for happiness and sadness, the arithmetical
calculation of \PB{\\{del}} shown here will produce \PB{\T{\^100}} if a move in
direction~\PB{\|d} is currently sad, \PB{\T{\^001}} if that move is currently
happy.

\Y\B\4\D$\\{code}(\|r,\|c,\|d,\|p)$ \5
$(((((\|r\LL\T{2})+\|c)\LL\T{2})+\|d)\LL\T{2})+{}$\|p\par
\B\4\D$\\{tailcode}(\|r,\|c,\|d)$ \5
$\T{\^100}+(\\{code}(\T{0},\39\|r,\39\|c,\39\|d){}$)\par
\B\4\D$\\{bottom}$ \5
$\\{code}(\T{0},\39\T{0},\39\T{1},\39\T{1}{}$)\C{ an otherwise unused code }\par
\B\4\D$\\{bord}(\|r,\|c)$ \5
$\\{board}[((\|r)\LL\T{2})+(\|c){}$]\par
\B\4\D$\\{east}(\|r,\|c)$ \5
$\\{piece}\K\\{bord}(\|r,\39\|c+\T{1});$ $\\{del}\K(((\|c-\\{col}(\\{piece}))%
\GG\T{2})\AND\T{\^ff})+{}$\T{1}\par
\B\4\D$\\{west}(\|r,\|c)$ \5
$\\{piece}\K\\{bord}(\|r,\39\|c-\T{1});$ $\\{del}\K(((\\{col}(\\{piece})-\|c)%
\GG\T{2})\AND\T{\^ff})+{}$\T{1}\par
\B\4\D$\\{north}(\|r,\|c)$ \5
$\\{piece}\K\\{bord}(\|r-\T{1},\39\|c);$ $\\{del}\K(((\\{row}(\\{piece})-\|r)%
\GG\T{2})\AND\T{\^ff})+{}$\T{1}\par
\B\4\D$\\{south}(\|r,\|c)$ \5
$\\{piece}\K\\{bord}(\|r+\T{1},\39\|c);$ $\\{del}\K(((\|r-\\{row}(\\{piece}))%
\GG\T{2})\AND\T{\^ff})+{}$\T{1}\par
\fi

\M{10}\B\X10:Cases that move east from column 0\X${}\E{}$\6
\4\&{case} \\{tailcode}${}(\T{0},\39\T{0},\39\T{0}){}$:\5
\&{case} \\{code}${}(\T{0},\39\T{0},\39\T{0},\39\T{3}){}$:\5
\\{r0c0d0p3}:\5
${}\\{east}(\T{0},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c0d3p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{0},\39\T{3},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c1d0p2};\6
\4\&{case} \\{code}${}(\T{1},\39\T{0},\39\T{0},\39\T{1}){}$:\5
\\{r1c0d0p1}:\5
${}\\{east}(\T{1},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c0d3p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{0},\39\T{3},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c1d1p2};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{0},\39\T{0}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{0},\39\T{0},\39\T{3}){}$:\5
\\{r1c0d0p3}:\5
${}\\{east}(\T{1},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c0d3p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{0},\39\T{3},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c1d1p2};\6
\4\&{case} \\{code}${}(\T{2},\39\T{0},\39\T{0},\39\T{1}){}$:\5
\\{r2c0d0p1}:\5
${}\\{east}(\T{2},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c0d3p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{0},\39\T{3},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c1d1p2};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{0},\39\T{0}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{0},\39\T{0},\39\T{3}){}$:\5
\\{r2c0d0p3}:\5
${}\\{east}(\T{2},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c0d3p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{0},\39\T{3},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c1d1p2};\6
\4\&{case} \\{tailcode}${}(\T{3},\39\T{0},\39\T{0}){}$:\5
\&{case} \\{code}${}(\T{3},\39\T{0},\39\T{0},\39\T{1}){}$:\5
\\{r3c0d0p1}:\5
${}\\{east}(\T{3},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c0d1p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{0},\39\T{1},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c1d1p2};\par
\U8.\fi

\M{11}\B\X11:Cases that move east from column 1\X${}\E{}$\6
\4\&{case} \\{code}${}(\T{0},\39\T{1},\39\T{0},\39\T{2}){}$:\5
\\{r0c1d0p2}:\5
${}\\{east}(\T{0},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c1d3p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{1},\39\T{3},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c2d0p2};\6
\4\&{case} \\{tailcode}${}(\T{0},\39\T{1},\39\T{0}){}$:\5
\&{case} \\{code}${}(\T{0},\39\T{1},\39\T{0},\39\T{3}){}$:\5
\\{r0c1d0p3}:\5
${}\\{east}(\T{0},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c1d3p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{1},\39\T{3},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c2d0p2};\6
\4\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{0},\39\T{1}){}$:\5
\\{r1c1d0p1}:\5
${}\\{east}(\T{1},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c1d3p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{1},\39\T{3},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c2d1p2};\6
\4\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{0},\39\T{2}){}$:\5
\\{r1c1d0p2}:\5
${}\\{east}(\T{1},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c1d3p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{1},\39\T{3},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c2d1p2};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{1},\39\T{0}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{0},\39\T{3}){}$:\5
\\{r1c1d0p3}:\5
${}\\{east}(\T{1},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c1d3p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{1},\39\T{3},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c2d1p2};\6
\4\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{0},\39\T{1}){}$:\5
\\{r2c1d0p1}:\5
${}\\{east}(\T{2},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c1d3p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{1},\39\T{3},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c2d1p2};\6
\4\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{0},\39\T{2}){}$:\5
\\{r2c1d0p2}:\5
${}\\{east}(\T{2},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c1d3p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{1},\39\T{3},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c2d1p2};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{1},\39\T{0}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{0},\39\T{3}){}$:\5
\\{r2c1d0p3}:\5
${}\\{east}(\T{2},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c1d3p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{1},\39\T{3},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c2d1p2};\6
\4\&{case} \\{code}${}(\T{3},\39\T{1},\39\T{0},\39\T{1}){}$:\5
\\{r3c1d0p1}:\5
${}\\{east}(\T{3},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c1d2p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{1},\39\T{2},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c2d1p2};\6
\4\&{case} \\{tailcode}${}(\T{3},\39\T{1},\39\T{0}){}$:\5
\&{case} \\{code}${}(\T{3},\39\T{1},\39\T{0},\39\T{2}){}$:\5
\\{r3c1d0p2}:\5
${}\\{east}(\T{3},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c1d2p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{1},\39\T{2},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c2d1p2};\par
\U8.\fi

\M{12}\B\X12:Cases that move east from column 2\X${}\E{}$\6
\4\&{case} \\{code}${}(\T{0},\39\T{2},\39\T{0},\39\T{2}){}$:\5
\\{r0c2d0p2}:\5
${}\\{east}(\T{0},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c2d3p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{2},\39\T{3},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c3d3p2};\6
\4\&{case} \\{tailcode}${}(\T{0},\39\T{2},\39\T{0}){}$:\5
\&{case} \\{code}${}(\T{0},\39\T{2},\39\T{0},\39\T{3}){}$:\5
\\{r0c2d0p3}:\5
${}\\{east}(\T{0},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c2d3p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{2},\39\T{3},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c3d3p2};\6
\4\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{0},\39\T{1}){}$:\5
\\{r1c2d0p1}:\5
${}\\{east}(\T{1},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c2d3p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{2},\39\T{3},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c3d1p2};\6
\4\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{0},\39\T{2}){}$:\5
\\{r1c2d0p2}:\5
${}\\{east}(\T{1},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c2d3p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{2},\39\T{3},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c3d1p2};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{2},\39\T{0}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{0},\39\T{3}){}$:\5
\\{r1c2d0p3}:\5
${}\\{east}(\T{1},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c2d3p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{2},\39\T{3},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c3d1p2};\6
\4\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{0},\39\T{1}){}$:\5
\\{r2c2d0p1}:\5
${}\\{east}(\T{2},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c2d3p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{2},\39\T{3},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c3d1p2};\6
\4\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{0},\39\T{2}){}$:\5
\\{r2c2d0p2}:\5
${}\\{east}(\T{2},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c2d3p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{2},\39\T{3},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c3d1p2};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{2},\39\T{0}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{0},\39\T{3}){}$:\5
\\{r2c2d0p3}:\5
${}\\{east}(\T{2},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c2d3p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{2},\39\T{3},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c3d1p2};\6
\4\&{case} \\{code}${}(\T{3},\39\T{2},\39\T{0},\39\T{1}){}$:\5
\\{r3c2d0p1}:\5
${}\\{east}(\T{3},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c2d2p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{2},\39\T{2},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c3d1p2};\6
\4\&{case} \\{tailcode}${}(\T{3},\39\T{2},\39\T{0}){}$:\5
\&{case} \\{code}${}(\T{3},\39\T{2},\39\T{0},\39\T{2}){}$:\5
\\{r3c2d0p2}:\5
${}\\{east}(\T{3},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c2d2p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{2},\39\T{2},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c3d1p2};\par
\U8.\fi

\M{13}\B\X13:Cases that move west from column 1\X${}\E{}$\6
\4\&{case} \\{tailcode}${}(\T{0},\39\T{1},\39\T{2}){}$:\5
\&{case} \\{code}${}(\T{0},\39\T{1},\39\T{2},\39\T{0}){}$:\5
\\{r0c1d2p0}:\5
${}\\{west}(\T{0},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c1d0p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{1},\39\T{0},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c0d3p0};\6
\4\&{case} \\{code}${}(\T{0},\39\T{1},\39\T{2},\39\T{3}){}$:\5
\\{r0c1d2p3}:\5
${}\\{west}(\T{0},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c1d0p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{1},\39\T{0},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c0d3p0};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{1},\39\T{2}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{2},\39\T{1}){}$:\5
\\{r1c1d2p1}:\5
${}\\{west}(\T{1},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c1d1p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{1},\39\T{1},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c0d3p0};\6
\4\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{2},\39\T{0}){}$:\5
\\{r1c1d2p0}:\5
${}\\{west}(\T{1},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c1d1p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{1},\39\T{1},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c0d3p0};\6
\4\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{2},\39\T{3}){}$:\5
\\{r1c1d2p3}:\5
${}\\{west}(\T{1},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c1d1p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{1},\39\T{1},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c0d3p0};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{1},\39\T{2}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{2},\39\T{1}){}$:\5
\\{r2c1d2p1}:\5
${}\\{west}(\T{2},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c1d1p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{1},\39\T{1},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c0d3p0};\6
\4\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{2},\39\T{0}){}$:\5
\\{r2c1d2p0}:\5
${}\\{west}(\T{2},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c1d1p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{1},\39\T{1},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c0d3p0};\6
\4\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{2},\39\T{3}){}$:\5
\\{r2c1d2p3}:\5
${}\\{west}(\T{2},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c1d1p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{1},\39\T{1},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c0d3p0};\6
\4\&{case} \\{tailcode}${}(\T{3},\39\T{1},\39\T{2}){}$:\5
\&{case} \\{code}${}(\T{3},\39\T{1},\39\T{2},\39\T{1}){}$:\5
\\{r3c1d2p1}:\5
${}\\{west}(\T{3},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c1d1p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{1},\39\T{1},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c0d1p0};\6
\4\&{case} \\{code}${}(\T{3},\39\T{1},\39\T{2},\39\T{0}){}$:\5
\\{r3c1d2p0}:\5
${}\\{west}(\T{3},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c1d1p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{1},\39\T{1},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c0d1p0};\par
\U8.\fi

\M{14}\B\X14:Cases that move west from column 2\X${}\E{}$\6
\4\&{case} \\{tailcode}${}(\T{0},\39\T{2},\39\T{2}){}$:\5
\&{case} \\{code}${}(\T{0},\39\T{2},\39\T{2},\39\T{0}){}$:\5
\\{r0c2d2p0}:\5
${}\\{west}(\T{0},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c2d0p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{2},\39\T{0},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c1d3p0};\6
\4\&{case} \\{code}${}(\T{0},\39\T{2},\39\T{2},\39\T{3}){}$:\5
\\{r0c2d2p3}:\5
${}\\{west}(\T{0},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c2d0p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{2},\39\T{0},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c1d3p0};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{2},\39\T{2}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{2},\39\T{1}){}$:\5
\\{r1c2d2p1}:\5
${}\\{west}(\T{1},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c2d1p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{2},\39\T{1},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c1d3p0};\6
\4\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{2},\39\T{0}){}$:\5
\\{r1c2d2p0}:\5
${}\\{west}(\T{1},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c2d1p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{2},\39\T{1},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c1d3p0};\6
\4\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{2},\39\T{3}){}$:\5
\\{r1c2d2p3}:\5
${}\\{west}(\T{1},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c2d1p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{2},\39\T{1},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c1d3p0};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{2},\39\T{2}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{2},\39\T{1}){}$:\5
\\{r2c2d2p1}:\5
${}\\{west}(\T{2},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c2d1p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{2},\39\T{1},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c1d3p0};\6
\4\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{2},\39\T{0}){}$:\5
\\{r2c2d2p0}:\5
${}\\{west}(\T{2},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c2d1p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{2},\39\T{1},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c1d3p0};\6
\4\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{2},\39\T{3}){}$:\5
\\{r2c2d2p3}:\5
${}\\{west}(\T{2},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c2d1p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{2},\39\T{1},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c1d3p0};\6
\4\&{case} \\{tailcode}${}(\T{3},\39\T{2},\39\T{2}){}$:\5
\&{case} \\{code}${}(\T{3},\39\T{2},\39\T{2},\39\T{1}){}$:\5
\\{r3c2d2p1}:\5
${}\\{west}(\T{3},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c2d1p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{2},\39\T{1},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c1d2p0};\6
\4\&{case} \\{code}${}(\T{3},\39\T{2},\39\T{2},\39\T{0}){}$:\5
\\{r3c2d2p0}:\5
${}\\{west}(\T{3},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c2d1p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{2},\39\T{1},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c1d2p0};\par
\U8.\fi

\M{15}\B\X15:Cases that move west from column 3\X${}\E{}$\6
\4\&{case} \\{tailcode}${}(\T{0},\39\T{3},\39\T{2}){}$:\5
\&{case} \\{code}${}(\T{0},\39\T{3},\39\T{2},\39\T{3}){}$:\5
\\{r0c3d2p3}:\5
${}\\{west}(\T{0},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c3d3p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{3},\39\T{3},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c2d3p0};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{3},\39\T{2}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{3},\39\T{2},\39\T{1}){}$:\5
\\{r1c3d2p1}:\5
${}\\{west}(\T{1},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c3d1p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{3},\39\T{1},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c2d3p0};\6
\4\&{case} \\{code}${}(\T{1},\39\T{3},\39\T{2},\39\T{3}){}$:\5
\\{r1c3d2p3}:\5
${}\\{west}(\T{1},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c3d1p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{3},\39\T{1},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c2d3p0};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{3},\39\T{2}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{3},\39\T{2},\39\T{1}){}$:\5
\\{r2c3d2p1}:\5
${}\\{west}(\T{2},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c3d1p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{3},\39\T{1},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c2d3p0};\6
\4\&{case} \\{code}${}(\T{2},\39\T{3},\39\T{2},\39\T{3}){}$:\5
\\{r2c3d2p3}:\5
${}\\{west}(\T{2},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c3d1p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{3},\39\T{1},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c2d3p0};\6
\4\&{case} \\{tailcode}${}(\T{3},\39\T{3},\39\T{2}){}$:\5
\&{case} \\{code}${}(\T{3},\39\T{3},\39\T{2},\39\T{1}){}$:\5
\\{r3c3d2p1}:\5
${}\\{west}(\T{3},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c3d1p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{3},\39\T{1},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c2d2p0};\par
\U8.\fi

\M{16}\B\X16:Cases that move north from row 1\X${}\E{}$\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{0},\39\T{1}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{0},\39\T{1},\39\T{0}){}$:\5
\\{r1c0d1p0}:\5
${}\\{north}(\T{1},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c0d0p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{0},\39\T{0},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c0d0p3};\6
\4\&{case} \\{code}${}(\T{1},\39\T{0},\39\T{1},\39\T{3}){}$:\5
\\{r1c0d1p3}:\5
${}\\{north}(\T{1},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c0d0p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{0},\39\T{0},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c0d0p3};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{1},\39\T{1}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{1},\39\T{0}){}$:\5
\\{r1c1d1p0}:\5
${}\\{north}(\T{1},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c1d0p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{1},\39\T{0},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c1d2p3};\6
\4\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{1},\39\T{2}){}$:\5
\\{r1c1d1p2}:\5
${}\\{north}(\T{1},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c1d0p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{1},\39\T{0},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c1d2p3};\6
\4\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{1},\39\T{3}){}$:\5
\\{r1c1d1p3}:\5
${}\\{north}(\T{1},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c1d0p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{1},\39\T{0},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c1d2p3};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{2},\39\T{1}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{1},\39\T{0}){}$:\5
\\{r1c2d1p0}:\5
${}\\{north}(\T{1},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c2d0p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{2},\39\T{0},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c2d2p3};\6
\4\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{1},\39\T{2}){}$:\5
\\{r1c2d1p2}:\5
${}\\{north}(\T{1},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c2d0p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{2},\39\T{0},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c2d2p3};\6
\4\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{1},\39\T{3}){}$:\5
\\{r1c2d1p3}:\5
${}\\{north}(\T{1},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c2d0p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{2},\39\T{0},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c2d2p3};\6
\4\&{case} \\{code}${}(\T{1},\39\T{3},\39\T{1},\39\T{2}){}$:\5
\\{r1c3d1p2}:\5
${}\\{north}(\T{1},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c3d3p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{3},\39\T{3},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c3d2p3};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{3},\39\T{1}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{3},\39\T{1},\39\T{3}){}$:\5
\\{r1c3d1p3}:\5
${}\\{north}(\T{1},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c3d3p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{3},\39\T{3},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r0c3d2p3};\par
\U8.\fi

\M{17}\B\X17:Cases that move north from row 2\X${}\E{}$\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{0},\39\T{1}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{0},\39\T{1},\39\T{0}){}$:\5
\\{r2c0d1p0}:\5
${}\\{north}(\T{2},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c0d0p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{0},\39\T{0},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c0d1p3};\6
\4\&{case} \\{code}${}(\T{2},\39\T{0},\39\T{1},\39\T{3}){}$:\5
\\{r2c0d1p3}:\5
${}\\{north}(\T{2},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c0d0p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{0},\39\T{0},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c0d1p3};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{1},\39\T{1}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{1},\39\T{0}){}$:\5
\\{r2c1d1p0}:\5
${}\\{north}(\T{2},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c1d0p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{1},\39\T{0},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c1d2p3};\6
\4\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{1},\39\T{2}){}$:\5
\\{r2c1d1p2}:\5
${}\\{north}(\T{2},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c1d0p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{1},\39\T{0},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c1d2p3};\6
\4\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{1},\39\T{3}){}$:\5
\\{r2c1d1p3}:\5
${}\\{north}(\T{2},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c1d0p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{1},\39\T{0},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c1d2p3};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{2},\39\T{1}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{1},\39\T{0}){}$:\5
\\{r2c2d1p0}:\5
${}\\{north}(\T{2},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c2d0p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{2},\39\T{0},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c2d2p3};\6
\4\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{1},\39\T{2}){}$:\5
\\{r2c2d1p2}:\5
${}\\{north}(\T{2},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c2d0p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{2},\39\T{0},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c2d2p3};\6
\4\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{1},\39\T{3}){}$:\5
\\{r2c2d1p3}:\5
${}\\{north}(\T{2},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c2d0p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{2},\39\T{0},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c2d2p3};\6
\4\&{case} \\{code}${}(\T{2},\39\T{3},\39\T{1},\39\T{2}){}$:\5
\\{r2c3d1p2}:\5
${}\\{north}(\T{2},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c3d3p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{3},\39\T{3},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c3d2p3};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{3},\39\T{1}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{3},\39\T{1},\39\T{3}){}$:\5
\\{r2c3d1p3}:\5
${}\\{north}(\T{2},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c3d3p3};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{3},\39\T{3},\39\T{3}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c3d2p3};\par
\U8.\fi

\M{18}\B\X18:Cases that move north from row 3\X${}\E{}$\6
\4\&{case} \\{tailcode}${}(\T{3},\39\T{0},\39\T{1}){}$:\5
\&{case} \\{code}${}(\T{3},\39\T{0},\39\T{1},\39\T{0}){}$:\5
\\{r3c0d1p0}:\5
${}\\{north}(\T{3},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c0d0p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{0},\39\T{0},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c0d1p3};\6
\4\&{case} \\{tailcode}${}(\T{3},\39\T{1},\39\T{1}){}$:\5
\&{case} \\{code}${}(\T{3},\39\T{1},\39\T{1},\39\T{0}){}$:\5
\\{r3c1d1p0}:\5
${}\\{north}(\T{3},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c1d0p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{1},\39\T{0},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c1d2p3};\6
\4\&{case} \\{code}${}(\T{3},\39\T{1},\39\T{1},\39\T{2}){}$:\5
\\{r3c1d1p2}:\5
${}\\{north}(\T{3},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c1d0p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{1},\39\T{0},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c1d2p3};\6
\4\&{case} \\{tailcode}${}(\T{3},\39\T{2},\39\T{1}){}$:\5
\&{case} \\{code}${}(\T{3},\39\T{2},\39\T{1},\39\T{0}){}$:\5
\\{r3c2d1p0}:\5
${}\\{north}(\T{3},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c2d0p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{2},\39\T{0},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c2d2p3};\6
\4\&{case} \\{code}${}(\T{3},\39\T{2},\39\T{1},\39\T{2}){}$:\5
\\{r3c2d1p2}:\5
${}\\{north}(\T{3},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c2d0p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{2},\39\T{0},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c2d2p3};\6
\4\&{case} \\{tailcode}${}(\T{3},\39\T{3},\39\T{1}){}$:\5
\&{case} \\{code}${}(\T{3},\39\T{3},\39\T{1},\39\T{2}){}$:\5
\\{r3c3d1p2}:\5
${}\\{north}(\T{3},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r3c3d2p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{3},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{3},\39\T{3},\39\T{2},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c3d2p3};\par
\U8.\fi

\M{19}\B\X19:Cases that move south from row 0\X${}\E{}$\6
\4\&{case} \\{tailcode}${}(\T{0},\39\T{0},\39\T{3}){}$:\5
\&{case} \\{code}${}(\T{0},\39\T{0},\39\T{3},\39\T{0}){}$:\5
\\{r0c0d3p0}:\5
${}\\{south}(\T{0},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c0d0p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{0},\39\T{0},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c0d0p1};\6
\4\&{case} \\{code}${}(\T{0},\39\T{1},\39\T{3},\39\T{0}){}$:\5
\\{r0c1d3p0}:\5
${}\\{south}(\T{0},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c1d2p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{1},\39\T{2},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c1d0p1};\6
\4\&{case} \\{tailcode}${}(\T{0},\39\T{1},\39\T{3}){}$:\5
\&{case} \\{code}${}(\T{0},\39\T{1},\39\T{3},\39\T{2}){}$:\5
\\{r0c1d3p2}:\5
${}\\{south}(\T{0},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c1d2p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{1},\39\T{2},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c1d0p1};\6
\4\&{case} \\{code}${}(\T{0},\39\T{2},\39\T{3},\39\T{0}){}$:\5
\\{r0c2d3p0}:\5
${}\\{south}(\T{0},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c2d2p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{2},\39\T{2},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c2d0p1};\6
\4\&{case} \\{tailcode}${}(\T{0},\39\T{2},\39\T{3}){}$:\5
\&{case} \\{code}${}(\T{0},\39\T{2},\39\T{3},\39\T{2}){}$:\5
\\{r0c2d3p2}:\5
${}\\{south}(\T{0},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c2d2p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{2},\39\T{2},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c2d0p1};\6
\4\&{case} \\{tailcode}${}(\T{0},\39\T{3},\39\T{3}){}$:\5
\&{case} \\{code}${}(\T{0},\39\T{3},\39\T{3},\39\T{2}){}$:\5
\\{r0c3d3p2}:\5
${}\\{south}(\T{0},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r0c3d2p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{0},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{0},\39\T{3},\39\T{2},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r1c3d3p1};\par
\U8.\fi

\M{20}\B\X20:Cases that move south from row 1\X${}\E{}$\6
\4\&{case} \\{code}${}(\T{1},\39\T{0},\39\T{3},\39\T{0}){}$:\5
\\{r1c0d3p0}:\5
${}\\{south}(\T{1},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c0d1p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{0},\39\T{1},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c0d0p1};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{0},\39\T{3}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{0},\39\T{3},\39\T{1}){}$:\5
\\{r1c0d3p1}:\5
${}\\{south}(\T{1},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c0d1p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{0},\39\T{1},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c0d0p1};\6
\4\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{3},\39\T{0}){}$:\5
\\{r1c1d3p0}:\5
${}\\{south}(\T{1},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c1d2p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{1},\39\T{2},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c1d0p1};\6
\4\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{3},\39\T{1}){}$:\5
\\{r1c1d3p1}:\5
${}\\{south}(\T{1},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c1d2p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{1},\39\T{2},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c1d0p1};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{1},\39\T{3}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{3},\39\T{2}){}$:\5
\\{r1c1d3p2}:\5
${}\\{south}(\T{1},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c1d2p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{1},\39\T{2},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c1d0p1};\6
\4\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{3},\39\T{0}){}$:\5
\\{r1c2d3p0}:\5
${}\\{south}(\T{1},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c2d2p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{2},\39\T{2},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c2d0p1};\6
\4\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{3},\39\T{1}){}$:\5
\\{r1c2d3p1}:\5
${}\\{south}(\T{1},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c2d2p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{2},\39\T{2},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c2d0p1};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{2},\39\T{3}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{3},\39\T{2}){}$:\5
\\{r1c2d3p2}:\5
${}\\{south}(\T{1},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c2d2p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{2},\39\T{2},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c2d0p1};\6
\4\&{case} \\{code}${}(\T{1},\39\T{3},\39\T{3},\39\T{1}){}$:\5
\\{r1c3d3p1}:\5
${}\\{south}(\T{1},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c3d2p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{3},\39\T{2},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c3d3p1};\6
\4\&{case} \\{tailcode}${}(\T{1},\39\T{3},\39\T{3}){}$:\5
\&{case} \\{code}${}(\T{1},\39\T{3},\39\T{3},\39\T{2}){}$:\5
\\{r1c3d3p2}:\5
${}\\{south}(\T{1},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r1c3d2p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{1},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{1},\39\T{3},\39\T{2},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r2c3d3p1};\par
\U8.\fi

\M{21}Yes, this has been boring. But now we're in the last such section.

\Y\B\4\X21:Cases that move south from row 2\X${}\E{}$\6
\4\&{case} \\{code}${}(\T{2},\39\T{0},\39\T{3},\39\T{0}){}$:\5
\\{r2c0d3p0}:\5
${}\\{south}(\T{2},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c0d1p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{0},\39\T{1},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c0d0p1};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{0},\39\T{3}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{0},\39\T{3},\39\T{1}){}$:\5
\\{r2c0d3p1}:\5
${}\\{south}(\T{2},\39\T{0});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c0d1p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{0})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{0},\39\T{1},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c0d0p1};\6
\4\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{3},\39\T{0}){}$:\5
\\{r2c1d3p0}:\5
${}\\{south}(\T{2},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c1d2p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{1},\39\T{2},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c1d0p1};\6
\4\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{3},\39\T{1}){}$:\5
\\{r2c1d3p1}:\5
${}\\{south}(\T{2},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c1d2p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{1},\39\T{2},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c1d0p1};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{1},\39\T{3}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{3},\39\T{2}){}$:\5
\\{r2c1d3p2}:\5
${}\\{south}(\T{2},\39\T{1});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c1d2p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{1})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{1},\39\T{2},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c1d0p1};\6
\4\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{3},\39\T{0}){}$:\5
\\{r2c2d3p0}:\5
${}\\{south}(\T{2},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c2d2p0};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{2},\39\T{2},\39\T{0}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c2d0p1};\6
\4\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{3},\39\T{1}){}$:\5
\\{r2c2d3p1}:\5
${}\\{south}(\T{2},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c2d2p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{2},\39\T{2},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c2d0p1};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{2},\39\T{3}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{3},\39\T{2}){}$:\5
\\{r2c2d3p2}:\5
${}\\{south}(\T{2},\39\T{2});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c2d2p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{2})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{2},\39\T{2},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c2d0p1};\6
\4\&{case} \\{code}${}(\T{2},\39\T{3},\39\T{3},\39\T{1}){}$:\5
\\{r2c3d3p1}:\5
${}\\{south}(\T{2},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c3d2p1};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{3},\39\T{2},\39\T{1}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c3d2p1};\6
\4\&{case} \\{tailcode}${}(\T{2},\39\T{3},\39\T{3}){}$:\5
\&{case} \\{code}${}(\T{2},\39\T{3},\39\T{3},\39\T{2}){}$:\5
\\{r2c3d3p2}:\5
${}\\{south}(\T{2},\39\T{3});{}$\6
\&{if} ${}(\|t\Z\\{del}){}$\5
${}\{{}$\5
\1\&{if} ${}(\|t\E\\{del}){}$\1\5
\&{goto} \\{win};\5
\2\&{else}\1\5
\&{goto} \\{r2c3d2p2};\5
\2${}\}{}$\2\6
${}\\{bord}(\T{2},\39\T{3})\K\\{piece},\39\\{stack}[\|s\PP]\K(\|t\LL\T{16})+%
\\{code}(\T{2},\39\T{3},\39\T{2},\39\T{2}),\39\|t\MRL{-{\K}}\\{del};{}$\6
\&{goto} \\{r3c3d2p1};\par
\U8.\fi

\M{22}The cases $(r,c,d,p)$ with $d=p$ represent the times when we
are backtracking and must restore the previous board position.

\Y\B\4\X22:Cases that move back\X${}\E{}$\6
\4\&{case} \\{code}${}(\T{0},\39\T{0},\39\T{0},\39\T{0}){}$:\5
\\{r0c0d0p0}:\5
${}\\{bord}(\T{0},\39\T{0})\K\\{bord}(\T{0},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{0},\39\T{0},\39\T{3},\39\T{3}){}$:\5
\\{r0c0d3p3}:\5
${}\\{bord}(\T{0},\39\T{0})\K\\{bord}(\T{1},\39\T{0}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{0},\39\T{1},\39\T{0},\39\T{0}){}$:\5
\\{r0c1d0p0}:\5
${}\\{bord}(\T{0},\39\T{1})\K\\{bord}(\T{0},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{0},\39\T{1},\39\T{2},\39\T{2}){}$:\5
\\{r0c1d2p2}:\5
${}\\{bord}(\T{0},\39\T{1})\K\\{bord}(\T{0},\39\T{0}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{0},\39\T{1},\39\T{3},\39\T{3}){}$:\5
\\{r0c1d3p3}:\5
${}\\{bord}(\T{0},\39\T{1})\K\\{bord}(\T{1},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{0},\39\T{2},\39\T{0},\39\T{0}){}$:\5
\\{r0c2d0p0}:\5
${}\\{bord}(\T{0},\39\T{2})\K\\{bord}(\T{0},\39\T{3}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{0},\39\T{2},\39\T{2},\39\T{2}){}$:\5
\\{r0c2d2p2}:\5
${}\\{bord}(\T{0},\39\T{2})\K\\{bord}(\T{0},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{0},\39\T{2},\39\T{3},\39\T{3}){}$:\5
\\{r0c2d3p3}:\5
${}\\{bord}(\T{0},\39\T{2})\K\\{bord}(\T{1},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{0},\39\T{3},\39\T{2},\39\T{2}){}$:\5
\\{r0c3d2p2}:\5
${}\\{bord}(\T{0},\39\T{3})\K\\{bord}(\T{0},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{0},\39\T{3},\39\T{3},\39\T{3}){}$:\5
\\{r0c3d3p3}:\5
${}\\{bord}(\T{0},\39\T{3})\K\\{bord}(\T{1},\39\T{3}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{0},\39\T{0},\39\T{0}){}$:\5
\\{r1c0d0p0}:\5
${}\\{bord}(\T{1},\39\T{0})\K\\{bord}(\T{1},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{0},\39\T{1},\39\T{1}){}$:\5
\\{r1c0d1p1}:\5
${}\\{bord}(\T{1},\39\T{0})\K\\{bord}(\T{0},\39\T{0}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{0},\39\T{3},\39\T{3}){}$:\5
\\{r1c0d3p3}:\5
${}\\{bord}(\T{1},\39\T{0})\K\\{bord}(\T{2},\39\T{0}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{0},\39\T{0}){}$:\5
\\{r1c1d0p0}:\5
${}\\{bord}(\T{1},\39\T{1})\K\\{bord}(\T{1},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{1},\39\T{1}){}$:\5
\\{r1c1d1p1}:\5
${}\\{bord}(\T{1},\39\T{1})\K\\{bord}(\T{0},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{2},\39\T{2}){}$:\5
\\{r1c1d2p2}:\5
${}\\{bord}(\T{1},\39\T{1})\K\\{bord}(\T{1},\39\T{0}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{1},\39\T{3},\39\T{3}){}$:\5
\\{r1c1d3p3}:\5
${}\\{bord}(\T{1},\39\T{1})\K\\{bord}(\T{2},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{0},\39\T{0}){}$:\5
\\{r1c2d0p0}:\5
${}\\{bord}(\T{1},\39\T{2})\K\\{bord}(\T{1},\39\T{3}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{1},\39\T{1}){}$:\5
\\{r1c2d1p1}:\5
${}\\{bord}(\T{1},\39\T{2})\K\\{bord}(\T{0},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{2},\39\T{2}){}$:\5
\\{r1c2d2p2}:\5
${}\\{bord}(\T{1},\39\T{2})\K\\{bord}(\T{1},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{2},\39\T{3},\39\T{3}){}$:\5
\\{r1c2d3p3}:\5
${}\\{bord}(\T{1},\39\T{2})\K\\{bord}(\T{2},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{3},\39\T{1},\39\T{1}){}$:\5
\\{r1c3d1p1}:\5
${}\\{bord}(\T{1},\39\T{3})\K\\{bord}(\T{0},\39\T{3}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{3},\39\T{2},\39\T{2}){}$:\5
\\{r1c3d2p2}:\5
${}\\{bord}(\T{1},\39\T{3})\K\\{bord}(\T{1},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{1},\39\T{3},\39\T{3},\39\T{3}){}$:\5
\\{r1c3d3p3}:\5
${}\\{bord}(\T{1},\39\T{3})\K\\{bord}(\T{2},\39\T{3}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{0},\39\T{0},\39\T{0}){}$:\5
\\{r2c0d0p0}:\5
${}\\{bord}(\T{2},\39\T{0})\K\\{bord}(\T{2},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{0},\39\T{1},\39\T{1}){}$:\5
\\{r2c0d1p1}:\5
${}\\{bord}(\T{2},\39\T{0})\K\\{bord}(\T{1},\39\T{0}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{0},\39\T{3},\39\T{3}){}$:\5
\\{r2c0d3p3}:\5
${}\\{bord}(\T{2},\39\T{0})\K\\{bord}(\T{3},\39\T{0}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{0},\39\T{0}){}$:\5
\\{r2c1d0p0}:\5
${}\\{bord}(\T{2},\39\T{1})\K\\{bord}(\T{2},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{1},\39\T{1}){}$:\5
\\{r2c1d1p1}:\5
${}\\{bord}(\T{2},\39\T{1})\K\\{bord}(\T{1},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{2},\39\T{2}){}$:\5
\\{r2c1d2p2}:\5
${}\\{bord}(\T{2},\39\T{1})\K\\{bord}(\T{2},\39\T{0}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{1},\39\T{3},\39\T{3}){}$:\5
\\{r2c1d3p3}:\5
${}\\{bord}(\T{2},\39\T{1})\K\\{bord}(\T{3},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{0},\39\T{0}){}$:\5
\\{r2c2d0p0}:\5
${}\\{bord}(\T{2},\39\T{2})\K\\{bord}(\T{2},\39\T{3}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{1},\39\T{1}){}$:\5
\\{r2c2d1p1}:\5
${}\\{bord}(\T{2},\39\T{2})\K\\{bord}(\T{1},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{2},\39\T{2}){}$:\5
\\{r2c2d2p2}:\5
${}\\{bord}(\T{2},\39\T{2})\K\\{bord}(\T{2},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{2},\39\T{3},\39\T{3}){}$:\5
\\{r2c2d3p3}:\5
${}\\{bord}(\T{2},\39\T{2})\K\\{bord}(\T{3},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{3},\39\T{1},\39\T{1}){}$:\5
\\{r2c3d1p1}:\5
${}\\{bord}(\T{2},\39\T{3})\K\\{bord}(\T{1},\39\T{3}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{3},\39\T{2},\39\T{2}){}$:\5
\\{r2c3d2p2}:\5
${}\\{bord}(\T{2},\39\T{3})\K\\{bord}(\T{2},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{2},\39\T{3},\39\T{3},\39\T{3}){}$:\5
\\{r2c3d3p3}:\5
${}\\{bord}(\T{2},\39\T{3})\K\\{bord}(\T{3},\39\T{3}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{3},\39\T{0},\39\T{0},\39\T{0}){}$:\5
\\{r3c0d0p0}:\5
${}\\{bord}(\T{3},\39\T{0})\K\\{bord}(\T{3},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{3},\39\T{0},\39\T{1},\39\T{1}){}$:\5
\\{r3c0d1p1}:\5
${}\\{bord}(\T{3},\39\T{0})\K\\{bord}(\T{2},\39\T{0}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{3},\39\T{1},\39\T{0},\39\T{0}){}$:\5
\\{r3c1d0p0}:\5
${}\\{bord}(\T{3},\39\T{1})\K\\{bord}(\T{3},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{3},\39\T{1},\39\T{1},\39\T{1}){}$:\5
\\{r3c1d1p1}:\5
${}\\{bord}(\T{3},\39\T{1})\K\\{bord}(\T{2},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{3},\39\T{1},\39\T{2},\39\T{2}){}$:\5
\\{r3c1d2p2}:\5
${}\\{bord}(\T{3},\39\T{1})\K\\{bord}(\T{3},\39\T{0}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{3},\39\T{2},\39\T{0},\39\T{0}){}$:\5
\\{r3c2d0p0}:\5
${}\\{bord}(\T{3},\39\T{2})\K\\{bord}(\T{3},\39\T{3}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{3},\39\T{2},\39\T{1},\39\T{1}){}$:\5
\\{r3c2d1p1}:\5
${}\\{bord}(\T{3},\39\T{2})\K\\{bord}(\T{2},\39\T{2}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{3},\39\T{2},\39\T{2},\39\T{2}){}$:\5
\\{r3c2d2p2}:\5
${}\\{bord}(\T{3},\39\T{2})\K\\{bord}(\T{3},\39\T{1}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{3},\39\T{3},\39\T{1},\39\T{1}){}$:\5
\\{r3c3d1p1}:\5
${}\\{bord}(\T{3},\39\T{3})\K\\{bord}(\T{2},\39\T{3}){}$;\5
\&{goto} \\{switcher};\6
\4\&{case} \\{code}${}(\T{3},\39\T{3},\39\T{2},\39\T{2}){}$:\5
\\{r3c3d2p2}:\5
${}\\{bord}(\T{3},\39\T{3})\K\\{bord}(\T{3},\39\T{2}){}$;\5
\&{goto} \\{switcher};\par
\U8.\fi

\M{23}Great --- all the code has now been written for the inner loop.
Only two things remain: (1)~Getting the whole process rolling, and
(2)~eventually stopping the darn thing.

To get started, we need 2, 3, or 4 top-level (or should~I say
bottom-of-stack level) decisions about what direction to choose for the
first move. The \PB{\\{tailcode}}s were invented for precisely that purpose.

The process stops in one of two ways: Either there's no solution
(and we have try again with a higher quota), or a winning path was
found. It's easy to handle the first case without slowing anything
down, by simply putting the special code `\PB{\\{bottom}}' at the bottom of the
stack.

And I'll save the other case (the best case) for last.

\Y\B\4\X23:Try for a solution with \PB{\|t} more moves\X${}\E{}$\6
\&{for} ${}(\|j\K\T{0};{}$ ${}\|j<\T{16};{}$ ${}\|j\PP){}$\5
${}\{{}$\1\6
${}\\{board}[\|j]\K\\{start}[\|j];{}$\6
\&{if} ${}(\\{board}[\|j]<\T{0}){}$\1\5
${}\|k\K\|j;{}$\2\6
\4${}\}{}$\2\6
${}\\{stack}[\T{0}]\K(\|t\LL\T{16})+\\{bottom},\39\|s\K\T{1};{}$\6
\&{if} ${}(\\{col}(\|k)\I\T{3}){}$\1\5
${}\\{stack}[\|s\PP]\K(\|t\LL\T{16})+\\{tailcode}(\\{row}(\|k),\39\\{col}(\|k),%
\39\T{0}){}$;\C{ first move east }\2\6
\&{if} ${}(\\{row}(\|k)\I\T{0}){}$\1\5
${}\\{stack}[\|s\PP]\K(\|t\LL\T{16})+\\{tailcode}(\\{row}(\|k),\39\\{col}(\|k),%
\39\T{1}){}$;\C{ first move north }\2\6
\&{if} ${}(\\{col}(\|k)\I\T{0}){}$\1\5
${}\\{stack}[\|s\PP]\K(\|t\LL\T{16})+\\{tailcode}(\\{row}(\|k),\39\\{col}(\|k),%
\39\T{2}){}$;\C{ first move west }\2\6
\&{if} ${}(\\{row}(\|k)\I\T{3}){}$\1\5
${}\\{stack}[\|s\PP]\K(\|t\LL\T{16})+\\{tailcode}(\\{row}(\|k),\39\\{col}(\|k),%
\39\T{3}){}$;\C{ first move south }\2\6
\X8:Switch into action\X;\par
\U5.\fi

\M{24}Hey, we have a winner! The stack entries tell us how we got here,
so that we can easily give the user instructions about which piece
should be pushed at each step --- except that the last two steps haven't
been placed on the stack, because of my tricky optimizations.

At first I was going to leave those two steps as an exercise
for the user. Ha, ha, ha, a hilarious joke.
But, well, it looked a little too silly.
So I broke down and figured out how to deduce them from scratch.

\Y\B\4\X24:Output the results\X${}\E{}$\6
$\\{printf}(\.{"Solution\ in\ \%d+\%d\ m}\)\.{oves:\ "},\39\\{moves}\AND\T{%
\^ff},\39\\{moves}\GG\T{8});{}$\6
\&{if} ${}(\\{moves}>\T{1}){}$\5
${}\{{}$\1\6
\&{for} ${}(\|k\K\T{1};{}$ ${}(\\{stack}[\|k]\AND\T{\^ffff})\G\\{tailcode}(%
\T{0},\39\T{0},\39\T{0});{}$ ${}\|k\PP){}$\1\5
;\2\6
\&{for} ${}(\|j\K\T{0};{}$ ${}\|j<\T{16};{}$ ${}\|j\PP){}$\1\5
${}\\{board}[\|j]\K\\{start}[\|j];{}$\2\6
${}\|j\K(\\{stack}[\|k]\AND\T{\^ffff})\GG\T{4};{}$\6
\&{for} ${}(\|k\PP;{}$ ${}\|k<\|s;{}$ ${}\|k\PP){}$\5
${}\{{}$\1\6
${}\\{del}\K(\\{stack}[\|k]\AND\T{\^ffff})\GG\T{4};{}$\6
${}\\{printf}(\.{"\%x"},\39\\{board}[\\{del}]+\T{1});{}$\6
${}\\{board}[\|j]\K\\{board}[\\{del}],\39\|j\K\\{del};{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\|j\E\T{\^d}){}$\1\5
\\{printf}(\.{"ef\\n"});\2\6
\&{else} \&{if} ${}(\|j\E\T{\^7}){}$\1\5
\\{printf}(\.{"8c\\n"});\2\6
\&{else}\1\5
${}\\{printf}(\.{"b\%x\\n"},\39\\{bord}(\T{3},\39\T{3})+\T{1});{}$\2\6
\4${}\}{}$\5
\2\&{else}\5
${}\{{}$\1\6
\&{if} ${}(\\{moves}\E\T{1}){}$\1\5
${}\\{printf}(\.{"\%x"},\39\\{start}[\T{\^f}]+\T{1});{}$\2\6
${}\\{printf}(\.{"\ (\%d\ sec)\\n"},\39\\{time}(\T{0})-\\{timer});{}$\6
\4${}\}{}$\2\par
\U1.\fi

\N{1}{25}Index.
\fi

\inx
\fin
\con
