\input cwebmac
\hypertextrue\srcloctrue
\datethis

\N[2 achain2.w]{1}{1}Intro. This program is a revision of {\mc ACHAIN1}, which
you should
read first. I'm thinking that a few changes will speed that program up,
but as usual the proof of the pudding is in the eating.

The main changes here are:
(i)~If $a[j]=b[j]=2^j+2^k$, where $k<j-3$, one can
prove that $a[j-k-2]$ and $b[j-k-2]$ must both be $2^{j-k-2}$.
This additional constraint makes more chain values ``exist'' at
early stages.
(ii)~Every addition chain corresponds to a directed graph, and many different
addition chains can correspond to the same ``reduced'' digraph, as explained
at the end of Section 4.6.3. Therefore I've worked out a scheme by which
only one chain of each equivalence class is explored.

These changes, and the various changes that distinguish {\mc ACHAIN1}
from {\mc ACHAIN0}, are fairly independent. If I had time, I could therefore
experiment with various subsets, in order to see which of them are
really worth the effort. Who knows, maybe some of them actually
cause a slowdown, taken individually.

\Y\B\4\D$\\{nmax}$ \5
\T{10000000}\C{ should be less than $2^{24}$ on a 32-bit machine }\par
\Y\B\8\#\&{include} \.{<stdio.h>}\6
\8\#\&{include} \.{<stdlib.h>}\6
\8\#\&{include} \.{<time.h>}\6
\&{char} \|l[\\{nmax}];\6
\&{int} \|a[\T{128}]${},{}$ \|b[\T{128}];\6
\&{unsigned} \&{int} ${}\\{undo}[\T{128}*\T{128}];{}$\6
\&{int} \\{ptr};\C{ this many items of the \PB{\\{undo}} stack are in use }\6
\&{struct} ${}\{{}$\1\6
\&{int} \\{lbp}${},{}$ \\{ubp}${},{}$ \\{lbq}${},{}$ \\{ubq}${},{}$ \|r${},{}$ %
\\{ptrp}${},{}$ \\{ptrq};\2\6
${}\}{}$ \\{stack}[\T{128}];\6
\&{int} \\{tail}[\T{128}]${},{}$ \\{outdeg}[\T{128}]${},{}$ \\{outsum}[%
\T{128}]${},{}$ \\{limit}[\T{128}];\6
\&{FILE} ${}{*}\\{infile},{}$ ${}{*}\\{outfile};{}$\6
\&{int} \\{prime}[\T{1000}];\C{ 1000 primes will take us past 60 million }\6
\&{int} \\{pr};\C{ the number of primes known so far }\6
\&{char} \|x[\T{64}];\C{ exponents of the binary representation of $n$ }\7
\&{int} \\{main}(\&{int} \\{argc}${},\39{}$\&{char} ${}{*}\\{argv}[\,]){}$\1\1%
\2\2\6
${}\{{}$\1\6
\&{register} \&{int} \|i${},{}$ \|j${},{}$ \|n${},{}$ \|p${},{}$ \|q${},{}$ %
\|r${},{}$ \|s${},{}$ \\{ubp}${},{}$ \\{ubq}${},{}$ \\{lbp}${},{}$ %
\\{lbq}${},{}$ \\{ptrp}${},{}$ \\{ptrq};\6
\&{int} \\{lb}${},{}$ \\{ub}${},{}$ \\{timer}${}\K\T{0};{}$\7
\X2:Process the command line\X;\6
${}\\{prime}[\T{0}]\K\T{2},\39\\{pr}\K\T{1};{}$\6
${}\|a[\T{0}]\K\|b[\T{0}]\K\T{1},\39\|a[\T{1}]\K\|b[\T{1}]\K\T{2}{}$;\C{ an
addition chain always begins like this }\6
\&{for} ${}(\|n\K\T{1};{}$ ${}\|n<\\{nmax};{}$ ${}\|n\PP){}$\5
${}\{{}$\1\6
\X4:Input the next lower bound, \PB{\\{lb}}\X;\6
\X5:Find an upper bound; or in simple cases, set $l(n)$ and \PB{\&{goto} %
\\{done}}\X;\6
\X7:Backtrack until $l(n)$ is known\X;\6
\4\\{done}:\5
\X3:Output the value of $l(n)$\X;\6
\&{if} ${}(\|n\MOD\T{1000}\E\T{0}){}$\5
${}\{{}$\1\6
${}\|j\K\\{clock}(\,);{}$\6
${}\\{printf}(\.{"\%d..\%d\ done\ in\ \%.5g}\)\.{\ minutes\\n"},\39\|n-\T{999},%
\39\|n,\39(\&{double})(\|j-\\{timer})/(\T{60}*\.{CLOCKS\_PER\_SEC}));{}$\6
${}\\{timer}\K\|j;{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\par
\fi

\M[61 achain2.w]{2}\B\X2:Process the command line\X${}\E{}$\6
\&{if} ${}(\\{argc}\I\T{3}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Usage:\ \%s\ infile\ ou}\)\.{tfile\\n"},\39%
\\{argv}[\T{0}]);{}$\6
${}\\{exit}({-}\T{1});{}$\6
\4${}\}{}$\2\6
${}\\{infile}\K\\{fopen}(\\{argv}[\T{1}],\39\.{"r"});{}$\6
\&{if} ${}(\R\\{infile}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"I\ couldn't\ open\ `\%s}\)\.{'\ for\ reading!%
\\n"},\39\\{argv}[\T{1}]);{}$\6
${}\\{exit}({-}\T{2});{}$\6
\4${}\}{}$\2\6
${}\\{outfile}\K\\{fopen}(\\{argv}[\T{2}],\39\.{"w"});{}$\6
\&{if} ${}(\R\\{outfile}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"I\ couldn't\ open\ `\%s}\)\.{'\ for\ writing!%
\\n"},\39\\{argv}[\T{2}]);{}$\6
${}\\{exit}({-}\T{3});{}$\6
\4${}\}{}$\2\par
\U1.\fi

\M[77 achain2.w]{3}\B\X3:Output the value of $l(n)$\X${}\E{}$\6
$\\{fprintf}(\\{outfile},\39\.{"\%c"},\39\|l[\|n]+\.{'\ '});{}$\6
\\{fflush}(\\{outfile});\C{ make sure the result is viewable immediately }\par
\U1.\fi

\M[81 achain2.w]{4}At this point I compute the ``lower bound'' $\lfloor\lg n%
\rfloor+3$,
which is valid if $\nu n>4$. Simple cases where $\nu n\le 4$ will be
handled separately below.

\Y\B\4\X4:Input the next lower bound, \PB{\\{lb}}\X${}\E{}$\6
\&{for} ${}(\|q\K\|n,\39\|i\K{-}\T{1},\39\|j\K\T{0};{}$ \|q; ${}\|q\MRL{{\GG}{%
\K}}\T{1},\39\|i\PP){}$\1\6
\&{if} ${}(\|q\AND\T{1}){}$\1\5
${}\|x[\|j\PP]\K\|i{}$;\C{ now $i=\lfloor\lg n\rfloor$ and $j=\nu n$ }\2\2\6
${}\\{lb}\K\\{fgetc}(\\{infile})-\.{'\ '}{}$;\C{ \PB{\\{fgetc}} will return a
negative value after EOF }\6
\&{if} ${}(\\{lb}<\|i+\T{3}){}$\1\5
${}\\{lb}\K\|i+\T{3}{}$;\2\par
\U1.\fi

\M[92 achain2.w]{5}Three elementary and well-known upper bounds are considered:
(i)~$l(n)\le\lfloor\lg n\rfloor+\nu n-1$;
(ii)~$l(n)\le l(n-1)+1$;
(iii)~$l(n)\le l(p)+l(q)$ if $n=pq$.

Furthermore, there are four special cases
when Theorem 4.6.3C tells us we can save a step.
In this regard, I had to learn (the hard way) to avoid a
curious bug: Three of the four cases in Theorem 4.6.3C arise
when we factor $n$, so I thought I needed to test only the
other case here. But I got a surprise when $n=165$: Then
$n=3\cdot55$, so the factor method gave the upper bound
$l(3)+l(55)=10$; but another factorization, $n=5\cdot33$,
gives the better bound $l(5)+l(33)=9$.

\Y\B\4\X5:Find an upper bound; or in simple cases, set $l(n)$ and \PB{\&{goto} %
\\{done}}\X${}\E{}$\6
$\\{ub}\K\|i+\|j-\T{1};{}$\6
\&{if} ${}(\\{ub}>\|l[\|n-\T{1}]+\T{1}){}$\1\5
${}\\{ub}\K\|l[\|n-\T{1}]+\T{1};{}$\2\6
\X6:Try reducing \PB{\\{ub}} with the factor method\X;\6
${}\|l[\|n]\K\\{ub};{}$\6
\&{if} ${}(\|j\Z\T{3}){}$\1\5
\&{goto} \\{done};\2\6
\&{if} ${}(\|j\E\T{4}){}$\5
${}\{{}$\1\6
${}\|p\K\|x[\T{3}]-\|x[\T{2}],\39\|q\K\|x[\T{1}]-\|x[\T{0}];{}$\6
\&{if} ${}(\|p\E\|q\V\|p\E\|q+\T{1}\V(\|q\E\T{1}\W(\|p\E\T{3}\V(\|p\E\T{5}\W%
\|x[\T{2}]\E\|x[\T{1}]+\T{1})))){}$\1\5
${}\|l[\|n]\K\|i+\T{2};{}$\2\6
\&{goto} \\{done};\6
\4${}\}{}$\2\par
\U1.\fi

\M[119 achain2.w]{6}It's important to try the factor method even when \PB{$\|j%
\Z\T{4}$}, because
of the way prime numbers are recognized here: We would miss the
prime~3, for example.

On the other hand, we don't need to remember large primes that will
never arise as factors of any future~$n$.

\Y\B\4\X6:Try reducing \PB{\\{ub}} with the factor method\X${}\E{}$\6
\&{if} ${}(\|n>\T{2}){}$\1\6
\&{for} ${}(\|s\K\T{0};{}$  ; ${}\|s\PP){}$\5
${}\{{}$\1\6
${}\|p\K\\{prime}[\|s];{}$\6
${}\|q\K\|n/\|p;{}$\6
\&{if} ${}(\|n\E\|p*\|q){}$\5
${}\{{}$\1\6
\&{if} ${}(\|l[\|p]+\|l[\|q]<\\{ub}){}$\1\5
${}\\{ub}\K\|l[\|p]+\|l[\|q];{}$\2\6
\&{break};\6
\4${}\}{}$\2\6
\&{if} ${}(\|q\Z\|p){}$\5
${}\{{}$\C{ $n$ is prime }\1\6
\&{if} ${}(\\{pr}<\T{1000}){}$\1\5
${}\\{prime}[\\{pr}\PP]\K\|n;{}$\2\6
\&{break};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\2\par
\U5.\fi

\N[140 achain2.w]{1}{7}The interesting part.
Nontrivial changes begin to occur when we get into the guts of
the backtracking structure carried over from the previous versions of this
program, but the controlling loop at the outer level remains intact.

\Y\B\4\X7:Backtrack until $l(n)$ is known\X${}\E{}$\6
$\|l[\|n]\K\\{lb};{}$\6
\&{while} ${}(\\{lb}<\\{ub}){}$\5
${}\{{}$\1\6
\&{for} ${}(\|i\K\T{0};{}$ ${}\|i\Z\\{lb};{}$ ${}\|i\PP){}$\1\5
${}\\{outdeg}[\|i]\K\\{outsum}[\|i]\K\T{0};{}$\2\6
${}\|a[\\{lb}]\K\|b[\\{lb}]\K\|n;{}$\6
\&{for} ${}(\|i\K\T{2};{}$ ${}\|i<\\{lb};{}$ ${}\|i\PP){}$\1\5
${}\|a[\|i]\K\|a[\|i-\T{1}]+\T{1},\39\|b[\|i]\K\|b[\|i-\T{1}]\LL\T{1};{}$\2\6
\&{for} ${}(\|i\K\\{lb}-\T{1};{}$ ${}\|i\G\T{2};{}$ ${}\|i\MM){}$\5
${}\{{}$\1\6
\&{if} ${}((\|a[\|i]\LL\T{1})<\|a[\|i+\T{1}]){}$\1\5
${}\|a[\|i]\K(\|a[\|i+\T{1}]+\T{1})\GG\T{1};{}$\2\6
\&{if} ${}(\|b[\|i]\G\|b[\|i+\T{1}]){}$\1\5
${}\|b[\|i]\K\|b[\|i+\T{1}]-\T{1};{}$\2\6
\4${}\}{}$\2\6
\X9:Try to fix the rest of the chain; \PB{\&{goto} \\{done}} if it's possible%
\X;\6
${}\|l[\|n]\K\PP\\{lb};{}$\6
\4${}\}{}$\2\par
\U1.\fi

\M[159 achain2.w]{8}The main new idea implemented below is related to the
reduced digraph
representation of an addition chain, in which each element $a_s$
of the chain is effectively expressed as a sum of two or
more previous elements. For example, we might have $a_s=a_i+a_j+a_k+a_l$
where $i\le j\le k\le l$, represented by four arrows coming in to
node $a_s$ from nodes $a_i$, $a_j$, $a_k$, and $a_l$. The
colexicographically largest chain with this reduced digraph has
elements $a_k+a_l$, $a_j+a_k+a_l$, and $a_i+a_j+a_k+a_l$, and
we want to avoid redundant work by restricting consideration to such chains.

Fortunately there's an easy way to do this: For each chain element~$x$
we remember its ``tail,'' which is the smallest node that points to~$x$;
and we also keep track of the outdegree of each fixed element. If
$a_s=p+q$ with $p\ge q$, the tail of~$a_s$ is~$q$. And if $a_s$ has
outdegree~1 because it is used only as an input to $a_t$, we
insist that $q$ be at least as large as the tail of~$a_t$.

Array element \PB{\\{outdeg}[\|j]} is the current number of elements
$a_{s+1}$, $a_{s+2}$, \dots, that make use of $a_j$, and
array element \PB{\\{outsum}[\|j]} is the sum of all their subscripts.
These two arrays are easily updated as we move forward and backward
while backtracking; and \PB{\\{outsum}[\|j]} turns out to be exactly the
value~\PB{\|t} that we need to know when \PB{$\\{outdeg}[\|j]\K\T{1}$}.
(Has anybody seen this idea before? At my age I thought I had
seen all such simple tricks long ago, but maybe there still
are many more waiting to be discovered.)

One consequence: If \PB{$\\{outdeg}[\|s-\T{1}]\E\T{0}$}, we must either have
$a[s-1]={1\over2}a[s]$ or $a[s-1]\ge{2\over3}a[s]$. Because
after setting \PB{\|a[\|s]} to a value other than ${1\over2}a[s]$
we will have \PB{$\\{outdeg}[\|s-\T{1}]\K\T{1}$}, and
${1\over2}a[s-1]\ge\PB{\\{tail}}[s-1]\ge\PB{\\{tail}}[s]=a[s]-a[s-1]$.

\fi

\M[192 achain2.w]{9}We maintain a stack of subproblems, as usual when
backtracking.
Suppose \PB{\|a[\|t]} is the sum of two items already present, for all
$t>s$; we want to make sure that \PB{\|a[\|s]} is legitimate too.
For this purpose we try all combinations \PB{$\|a[\|s]\K\|p+\|q$} where
$p\ge a[s]/2$, trying to make both $p$ and $q$ present.
(By the nature of the algorithm, we'll have \PB{$\|a[\|s]\K\|b[\|s]$} at
the time we choose \PB{\|p} and~\PB{\|q}, because shorter addition
chains have been ruled out.)

As elements of \PB{\|a} and \PB{\|b} are changed, we record their previous
values on the \PB{\\{undo}} stack, so that we can easily restore them
later. Pointers \PB{\\{ptrp}} and \PB{\\{ptrq}} contain the limiting indexes
for undo information.

\Y\B\4\X9:Try to fix the rest of the chain; \PB{\&{goto} \\{done}} if it's
possible\X${}\E{}$\6
$\\{ptr}\K\T{0}{}$;\C{ clear the \PB{\\{undo}} stack }\6
\&{for} ${}(\|r\K\|s\K\\{lb};{}$ ${}\|s>\T{2};{}$ ${}\|s\MM){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{outdeg}[\|s]\E\T{1}){}$\1\5
${}\\{limit}[\|s]\K\\{tail}[\\{outsum}[\|s]]{}$;\5
\2\&{else}\1\5
${}\\{limit}[\|s]\K\T{1};{}$\2\6
\&{for} ( ; ${}\|r>\T{1}\W\|a[\|r-\T{1}]\E\|b[\|r-\T{1}];{}$ ${}\|r\MM){}$\1\5
;\2\6
\&{if} ${}(\\{outdeg}[\|s-\T{1}]\E\T{0}\W(\|a[\|s]\AND\T{1})){}$\1\5
${}\|q\K\|a[\|s]/\T{3}{}$;\5
\2\&{else}\1\5
${}\|q\K\|a[\|s]\GG\T{1};{}$\2\6
\&{for} ${}(\|p\K\|a[\|s]-\|q;{}$ ${}\|p\Z\|b[\|s-\T{1}];{}$ \,)\5
${}\{{}$\1\6
\&{if} ${}(\|p>\|b[\|r-\T{1}]){}$\5
${}\{{}$\1\6
\&{while} ${}(\|p>\|a[\|r]){}$\1\5
${}\|r\PP{}$;\C{ this step keeps \PB{$\|r<\|s$}, since \PB{$\|a[\|s-\T{1}]\K%
\|b[\|s-\T{1}]$} }\2\6
${}\|p\K\|a[\|r],\39\|q\K\|a[\|s]-\|p,\39\|r\PP;{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\|q<\\{limit}[\|s]){}$\1\5
\&{goto} \\{backup};\2\6
\X12:Find bounds $(\PB{\\{lbp}},\PB{\\{ubp}})$ and $(\PB{\\{lbq}},\PB{%
\\{ubq}})$ on where \PB{\|p} and \PB{\|q} can be inserted; but go to \PB{%
\\{failpq}} if they can't both be accommodated\X;\6
${}\\{ptrp}\K\\{ptr};{}$\6
\&{for} ( ; ${}\\{ubp}\G\\{lbp};{}$ ${}\\{ubp}\MM){}$\5
${}\{{}$\1\6
\X14:Put \PB{\|p} into the chain at location \PB{\\{ubp}}; \PB{\&{goto} %
\\{failp}} if there's a problem\X;\6
\&{if} ${}(\|p\E\|q){}$\1\5
\&{goto} \\{happiness};\2\6
\&{if} ${}(\\{ubq}\G\\{ubp}){}$\1\5
${}\\{ubq}\K\\{ubp}-\T{1};{}$\2\6
${}\\{ptrq}\K\\{ptr};{}$\6
\&{for} ( ; ${}\\{ubq}\G\\{lbq};{}$ ${}\\{ubq}\MM){}$\5
${}\{{}$\1\6
\X16:Put \PB{\|q} into the chain at location \PB{\\{ubq}}; \PB{\&{goto} %
\\{failq}} if there's a problem\X;\6
\4\\{happiness}:\5
\X10:Put local variables on the stack and update outdegrees\X;\6
\&{goto} \\{onward};\C{ now \PB{\|a[\|s]} is covered; try to fill in \PB{$\|a[%
\|s-\T{1}]$} }\6
\4\\{backup}:\5
${}\|s\PP;{}$\6
\&{if} ${}(\|s>\\{lb}){}$\1\5
\&{goto} \\{impossible};\2\6
\X11:Restore local variables from the stack and downdate outdegrees\X;\6
\&{if} ${}(\|p\E\|q){}$\1\5
\&{goto} \\{failp};\2\6
\4\\{failq}:\5
\&{while} ${}(\\{ptr}>\\{ptrq}){}$\1\5
\X13:Undo a change\X;\2\6
\4${}\}{}$\2\6
\4\\{failp}:\5
\&{while} ${}(\\{ptr}>\\{ptrp}){}$\1\5
\X13:Undo a change\X;\2\6
\4${}\}{}$\2\6
\4\\{failpq}:\5
\&{if} ${}(\|p\E\|q){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{outdeg}[\|s-\T{1}]\E\T{0}){}$\1\5
${}\|q\K\|a[\|s]/\T{3}+\T{1}{}$;\C{ will be decreased momentarily }\2\6
\&{if} ${}(\|q>\|b[\|s-\T{2}]){}$\1\5
${}\|q\K\|b[\|s-\T{2}];{}$\2\6
\&{else}\1\5
${}\|q\MM;{}$\2\6
${}\|p\K\|a[\|s]-\|q;{}$\6
\4${}\}{}$\5
\2\&{else}\1\5
${}\|p\PP,\39\|q\MM;{}$\2\6
\4${}\}{}$\2\6
\&{goto} \\{backup};\6
\4\\{onward}:\5
\&{continue};\6
\4${}\}{}$\2\6
\&{goto} \\{done};\6
\4\\{impossible}:\par
\U7.\fi

\M[254 achain2.w]{10}\B\X10:Put local variables on the stack and update
outdegrees\X${}\E{}$\6
$\\{tail}[\|s]\K\|q,\39\\{stack}[\|s].\|r\K\|r;{}$\6
${}\\{outdeg}[\\{ubp}]\PP,\39\\{outsum}[\\{ubp}]\MRL{+{\K}}\|s;{}$\6
${}\\{outdeg}[\\{ubq}]\PP,\39\\{outsum}[\\{ubq}]\MRL{+{\K}}\|s;{}$\6
${}\\{stack}[\|s].\\{lbp}\K\\{lbp},\39\\{stack}[\|s].\\{ubp}\K\\{ubp};{}$\6
${}\\{stack}[\|s].\\{lbq}\K\\{lbq},\39\\{stack}[\|s].\\{ubq}\K\\{ubq};{}$\6
${}\\{stack}[\|s].\\{ptrp}\K\\{ptrp},\39\\{stack}[\|s].\\{ptrq}\K\\{ptrq}{}$;%
\par
\U9.\fi

\M[262 achain2.w]{11}\B\X11:Restore local variables from the stack and downdate
outdegrees\X${}\E{}$\6
$\\{ptrq}\K\\{stack}[\|s].\\{ptrq},\39\\{ptrp}\K\\{stack}[\|s].\\{ptrp};{}$\6
${}\\{lbq}\K\\{stack}[\|s].\\{lbq},\39\\{ubq}\K\\{stack}[\|s].\\{ubq};{}$\6
${}\\{lbp}\K\\{stack}[\|s].\\{lbp},\39\\{ubp}\K\\{stack}[\|s].\\{ubp};{}$\6
${}\\{outdeg}[\\{ubq}]\MM,\39\\{outsum}[\\{ubq}]\MRL{-{\K}}\|s;{}$\6
${}\\{outdeg}[\\{ubp}]\MM,\39\\{outsum}[\\{ubp}]\MRL{-{\K}}\|s;{}$\6
${}\|q\K\\{tail}[\|s],\39\|p\K\|a[\|s]-\|q,\39\|r\K\\{stack}[\|s].\|r{}$;\par
\U9.\fi

\M[270 achain2.w]{12}After the test in this step is passed, we'll have \PB{$%
\\{ubp}>\\{ubq}$} and \PB{$\\{lbp}>\\{lbq}$}.

\Y\B\4\X12:Find bounds $(\PB{\\{lbp}},\PB{\\{ubp}})$ and $(\PB{\\{lbq}},\PB{%
\\{ubq}})$ on where \PB{\|p} and \PB{\|q} can be inserted; but go to \PB{%
\\{failpq}} if they can't both be accommodated\X${}\E{}$\6
$\\{lbp}\K\|l[\|p];{}$\6
\&{if} ${}(\\{lbp}\G\|s){}$\1\5
\&{goto} \\{failpq};\2\6
\&{while} ${}(\|b[\\{lbp}]<\|p){}$\1\5
${}\\{lbp}\PP;{}$\2\6
\&{if} ${}(\|a[\\{lbp}]>\|p){}$\1\5
\&{goto} \\{failpq};\2\6
\&{for} ${}(\\{ubp}\K\\{lbp};{}$ ${}\|a[\\{ubp}+\T{1}]\Z\|p;{}$ ${}\\{ubp}%
\PP){}$\1\5
;\2\6
\&{if} ${}(\\{ubp}\E\|s-\T{1}){}$\1\5
${}\\{lbp}\K\\{ubp};{}$\2\6
\&{if} ${}(\|p\E\|q){}$\1\5
${}\\{lbq}\K\\{lbp},\39\\{ubq}\K\\{ubp};{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{lbq}\K\|l[\|q];{}$\6
\&{if} ${}(\\{lbq}\G\\{ubp}){}$\1\5
\&{goto} \\{failpq};\2\6
\&{while} ${}(\|b[\\{lbq}]<\|q){}$\1\5
${}\\{lbq}\PP;{}$\2\6
\&{if} ${}(\\{lbq}\G\\{ubp}){}$\1\5
\&{goto} \\{failpq};\2\6
\&{if} ${}(\|a[\\{lbq}]>\|q){}$\1\5
\&{goto} \\{failpq};\2\6
\&{for} ${}(\\{ubq}\K\\{lbq};{}$ ${}\|a[\\{ubq}+\T{1}]\Z\|q\W\\{ubq}+\T{1}<%
\\{ubp};{}$ ${}\\{ubq}\PP){}$\1\5
;\2\6
\&{if} ${}(\\{lbp}\E\\{lbq}){}$\1\5
${}\\{lbp}\PP;{}$\2\6
\4${}\}{}$\2\par
\U9.\fi

\M[290 achain2.w]{13}The undoing mechanism is very simple: When changing \PB{%
\|a[\|j]}, we
put \PB{$(\|j\LL\T{24})+\|x$} on the \PB{\\{undo}} stack, where \PB{\|x} was
the former value.
Similarly, when changing \PB{\|b[\|j]}, we stack the value \PB{$(\T{1}\LL%
\T{31})+(\|j\LL\T{24})+\|x$}.

\Y\B\4\D$\\{newa}(\|j,\|y)$ \5
$\\{undo}[\\{ptr}\PP]\K(\|j\LL\T{24})+\|a[\|j],\39\|a[\|j]\K{}$\|y\par
\B\4\D$\\{newb}(\|j,\|y)$ \5
$\\{undo}[\\{ptr}\PP]\K(\T{1}\LL\T{31})+(\|j\LL\T{24})+\|b[\|j],\39\|b[\|j]%
\K{}$\|y\par
\Y\B\4\X13:Undo a change\X${}\E{}$\6
${}\{{}$\1\6
${}\|i\K\\{undo}[\MM\\{ptr}];{}$\6
\&{if} ${}(\|i\G\T{0}){}$\1\5
${}\|a[\|i\GG\T{24}]\K\|i\AND\T{\^ffffff};{}$\2\6
\&{else}\1\5
${}\|b[(\|i\AND\T{\^3fffffff})\GG\T{24}]\K\|i\AND\T{\^ffffff};{}$\2\6
\4${}\}{}$\2\par
\U9.\fi

\M[304 achain2.w]{14}At this point we know that $a[ubp]\le p\le b[ubp]$.

\Y\B\4\X14:Put \PB{\|p} into the chain at location \PB{\\{ubp}}; \PB{\&{goto} %
\\{failp}} if there's a problem\X${}\E{}$\6
\&{if} ${}(\|a[\\{ubp}]\I\|p){}$\5
${}\{{}$\1\6
${}\\{newa}(\\{ubp},\39\|p);{}$\6
\&{for} ${}(\|j\K\\{ubp}-\T{1};{}$ ${}(\|a[\|j]\LL\T{1})<\|a[\|j+\T{1}];{}$ ${}%
\|j\MM){}$\5
${}\{{}$\1\6
${}\|i\K(\|a[\|j+\T{1}]+\T{1})\GG\T{1};{}$\6
\&{if} ${}(\|i>\|b[\|j]){}$\1\5
\&{goto} \\{failp};\2\6
${}\\{newa}(\|j,\39\|i);{}$\6
\4${}\}{}$\2\6
\&{for} ${}(\|j\K\\{ubp}+\T{1};{}$ ${}\|a[\|j]\Z\|a[\|j-\T{1}];{}$ ${}\|j%
\PP){}$\5
${}\{{}$\1\6
${}\|i\K\|a[\|j-\T{1}]+\T{1};{}$\6
\&{if} ${}(\|i>\|b[\|j]){}$\1\5
\&{goto} \\{failp};\2\6
${}\\{newa}(\|j,\39\|i);{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\|b[\\{ubp}]\I\|p){}$\5
${}\{{}$\1\6
${}\\{newb}(\\{ubp},\39\|p);{}$\6
\&{for} ${}(\|j\K\\{ubp}-\T{1};{}$ ${}\|b[\|j]\G\|b[\|j+\T{1}];{}$ ${}\|j%
\MM){}$\5
${}\{{}$\1\6
${}\|i\K\|b[\|j+\T{1}]-\T{1};{}$\6
\&{if} ${}(\|i<\|a[\|j]){}$\1\5
\&{goto} \\{failp};\2\6
${}\\{newb}(\|j,\39\|i);{}$\6
\4${}\}{}$\2\6
\&{for} ${}(\|j\K\\{ubp}+\T{1};{}$ ${}\|b[\|j]>\|b[\|j-\T{1}]\LL\T{1};{}$ ${}%
\|j\PP){}$\5
${}\{{}$\1\6
${}\|i\K\|b[\|j-\T{1}]\LL\T{1};{}$\6
\&{if} ${}(\|i<\|a[\|j]){}$\1\5
\&{goto} \\{failp};\2\6
${}\\{newb}(\|j,\39\|i);{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\X15:Make forced moves if \PB{\|p} has a special form\X;\par
\U9.\fi

\M[335 achain2.w]{15}If, say, we've just set \PB{$\|a[\T{8}]\K\|b[\T{8}]\K%
\T{132}$}, special considerations apply,
because the only addition chains of length~8 for 132 are
$$\eqalign{
&1,2,4,8,16,32,64,128,132;\cr
&1,2,4,8,16,32,64,68,132;\cr
&1,2,4,8,16,32,64,66,132;\cr
&1,2,4,8,16,32,34,66,132;\cr
&1,2,4,8,16,32,33,66,132;\cr
&1,2,4,8,16,17,33,66,132.\cr}$$
The values of \PB{\|a[\T{4}]} and \PB{\|b[\T{4}]} must therefore be 16; and
then, of course,
we also must have \PB{$\|a[\T{3}]\K\|b[\T{3}]\K\T{8}$}, etc. Similar reasoning
applies
whenever we set $a[j]=b[j]=2^j+2^k$ for $k\le j-4$.

Such cases may seem extremely special. But my hunch is that they are
important, because efficient chains need such values. When we try
to prove that no efficient chain exists, we want to show that
such values can't be present. Numbers with small \PB{\|l[\|p]} are harder
to rule out, so it should be helpful to penalize them.

\Y\B\4\X15:Make forced moves if \PB{\|p} has a special form\X${}\E{}$\6
$\|i\K\|p-(\T{1}\LL(\\{ubp}-\T{1}));{}$\6
\&{if} ${}(\|i\W((\|i\AND(\|i-\T{1}))\E\T{0})\W(\|i\LL\T{4})<\|p){}$\5
${}\{{}$\1\6
\&{for} ${}(\|j\K\\{ubp}-\T{2};{}$ ${}(\|i\AND\T{1})\E\T{0};{}$ ${}\|i\MRL{{%
\GG}{\K}}\T{1},\39\|j\MM){}$\1\5
;\2\6
\&{if} ${}(\|b[\|j]<(\T{1}\LL\|j)){}$\1\5
\&{goto} \\{failp};\2\6
\&{for} ( ; ${}\|a[\|j]<(\T{1}\LL\|j);{}$ ${}\|j\MM){}$\1\5
${}\\{newa}(\|j,\39\T{1}\LL\|j);{}$\2\6
\4${}\}{}$\2\par
\U14.\fi

\M[362 achain2.w]{16}At this point we had better not assume that $a[ubq]\le q%
\le b[ubq]$,
because \PB{\|p} has just been inserted. That insertion can mess up the
bounds that we looked at when \PB{\\{lbq}} and \PB{\\{ubq}} were computed.

\Y\B\4\X16:Put \PB{\|q} into the chain at location \PB{\\{ubq}}; \PB{\&{goto} %
\\{failq}} if there's a problem\X${}\E{}$\6
\&{if} ${}(\|a[\\{ubq}]\I\|q){}$\5
${}\{{}$\1\6
\&{if} ${}(\|a[\\{ubq}]>\|q){}$\1\5
\&{goto} \\{failq};\2\6
${}\\{newa}(\\{ubq},\39\|q);{}$\6
\&{for} ${}(\|j\K\\{ubq}-\T{1};{}$ ${}(\|a[\|j]\LL\T{1})<\|a[\|j+\T{1}];{}$ ${}%
\|j\MM){}$\5
${}\{{}$\1\6
${}\|i\K(\|a[\|j+\T{1}]+\T{1})\GG\T{1};{}$\6
\&{if} ${}(\|i>\|b[\|j]){}$\1\5
\&{goto} \\{failq};\2\6
${}\\{newa}(\|j,\39\|i);{}$\6
\4${}\}{}$\2\6
\&{for} ${}(\|j\K\\{ubq}+\T{1};{}$ ${}\|a[\|j]\Z\|a[\|j-\T{1}];{}$ ${}\|j%
\PP){}$\5
${}\{{}$\1\6
${}\|i\K\|a[\|j-\T{1}]+\T{1};{}$\6
\&{if} ${}(\|i>\|b[\|j]){}$\1\5
\&{goto} \\{failq};\2\6
${}\\{newa}(\|j,\39\|i);{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\|b[\\{ubq}]\I\|q){}$\5
${}\{{}$\1\6
\&{if} ${}(\|b[\\{ubq}]<\|q){}$\1\5
\&{goto} \\{failq};\2\6
${}\\{newb}(\\{ubq},\39\|q);{}$\6
\&{for} ${}(\|j\K\\{ubq}-\T{1};{}$ ${}\|b[\|j]\G\|b[\|j+\T{1}];{}$ ${}\|j%
\MM){}$\5
${}\{{}$\1\6
${}\|i\K\|b[\|j+\T{1}]-\T{1};{}$\6
\&{if} ${}(\|i<\|a[\|j]){}$\1\5
\&{goto} \\{failq};\2\6
${}\\{newb}(\|j,\39\|i);{}$\6
\4${}\}{}$\2\6
\&{for} ${}(\|j\K\\{ubq}+\T{1};{}$ ${}\|b[\|j]>\|b[\|j-\T{1}]\LL\T{1};{}$ ${}%
\|j\PP){}$\5
${}\{{}$\1\6
${}\|i\K\|b[\|j-\T{1}]\LL\T{1};{}$\6
\&{if} ${}(\|i<\|a[\|j]){}$\1\5
\&{goto} \\{failq};\2\6
${}\\{newb}(\|j,\39\|i);{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\X17:Make forced moves if \PB{\|q} has a special form\X;\par
\U9.\fi

\M[397 achain2.w]{17}\B\X17:Make forced moves if \PB{\|q} has a special form%
\X${}\E{}$\6
$\|i\K\|q-(\T{1}\LL(\\{ubq}-\T{1}));{}$\6
\&{if} ${}(\|i\W((\|i\AND(\|i-\T{1}))\E\T{0})\W(\|i\LL\T{4})<\|q){}$\5
${}\{{}$\1\6
\&{for} ${}(\|j\K\\{ubq}-\T{2};{}$ ${}(\|i\AND\T{1})\E\T{0};{}$ ${}\|i\MRL{{%
\GG}{\K}}\T{1},\39\|j\MM){}$\1\5
;\2\6
\&{if} ${}(\|b[\|j]<(\T{1}\LL\|j)){}$\1\5
\&{goto} \\{failq};\2\6
\&{for} ( ; ${}\|a[\|j]<(\T{1}\LL\|j);{}$ ${}\|j\MM){}$\1\5
${}\\{newa}(\|j,\39\T{1}\LL\|j);{}$\2\6
\4${}\}{}$\2\par
\U16.\fi

\N[406 achain2.w]{1}{18}Index.


\fi


\inx
\fin
\con
