\input cwebmac
\srcloctrue

\N[1 commafree-eastman.w]{1}{1}Intro. This program is an iterative
implementation of
an interesting recursive algorithm due to
Willard~L. Eastman, {\sl IEEE Trans.\ \bf IT-11} (1965), 263--267:
Given a sequence of nonnegative integers $x=x_0x_1\ldots x_{n-1}$ of
odd length~$n$, where $x$ is not equal to any of its cyclic shifts
$x_k\ldots x_{n-1}x_0\ldots x_{k-1}$ for $1\le k<n$, we output a
cyclic shift $\sigma x$ such that the set of all such $\sigma x$
forms a commafree code of block length~$n$ (over an infinite alphabet).

The integers are given as command-line arguments.

The simplest nontrivial example occurs when $n=3$. If $x=abc$, where
$a$, $b$, and~$c$ aren't all equal, then exactly one of the cyclic shifts
$y_0y_1y_2=abc$, $bca$, $cab$ will satisfy $y_0>y_1\le y_2$, and we
choose that one. It's easy to check that the triples chosen in
this way are commafree.

Similar constructions are possible when $n=5$ or $n=7$. But the case
$n=9$ already gets a bit dicey, and when $n$ is really large it's not
at all clear that commafreeness is possible. Eastman's paper resolved
a conjecture made by Golomb, Gordon, and Welch in their pioneering paper about
comma-free codes (1958).

(Of course, it's not at all clear
that we would want to actually {\it use\/} a commafree code when
$n$ is large; but that's another story, and beside the point. The point
is that Eastman discovered a really interesting algorithm.)

\Y\B\4\D$\\{maxn}$ \5
\T{105}\par
\Y\B\8\#\&{include} \.{<stdio.h>}\6
\8\#\&{include} \.{<stdlib.h>}\6
\&{int} ${}\|x[\\{maxn}+\\{maxn}+\\{maxn}];{}$\6
\&{int} ${}\|b[\\{maxn}+\\{maxn}+\\{maxn}];{}$\6
\&{int} \\{bb}[\\{maxn}];\7
\X5:Subroutines\X;\7
\\{main}(\&{int} \\{argc}${},\39{}$\&{char} ${}{*}\\{argv}[\,]){}$\1\1\2\2\6
${}\{{}$\1\6
\&{register} \&{int} \|i${},{}$ \|j${},{}$ \|k${},{}$ \|n${},{}$ \|p${},{}$ %
\|q${},{}$ \|t${},{}$ \\{tt};\7
\X2:Process the command line\X;\6
\X3:Do Eastman's algorithm\X;\6
\4${}\}{}$\2\par
\fi

\M[44 commafree-eastman.w]{2}\B\X2:Process the command line\X${}\E{}$\6
\&{if} ${}(\\{argc}<\T{4}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Usage:\ \%s\ x1\ x2\ ...}\)\.{\ xn\\n"},\39%
\\{argv}[\T{0}]);{}$\6
${}\\{exit}({-}\T{1});{}$\6
\4${}\}{}$\2\6
${}\|n\K\\{argc}-\T{1};{}$\6
\&{if} ${}((\|n\AND\T{1})\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"The\ number\ of\ items}\)\.{,\ n,\ should\ be\
odd,\ }\)\.{not\ \%d!\\n"},\39\|n);{}$\6
${}\\{exit}({-}\T{2});{}$\6
\4${}\}{}$\2\6
\&{for} ${}(\|j\K\T{1};{}$ ${}\|j<\\{argc};{}$ ${}\|j\PP){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{sscanf}(\\{argv}[\|j],\39\.{"\%d"},\39{\AND}\|x[\|j-\T{1}])\I%
\T{1}\V\|x[\|j-\T{1}]<\T{0}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Argument\ \%d\ should\ }\)\.{be\ a\
nonnegative\ int}\)\.{eger,\ not\ `\%s'!\\n"},\39\|j,\39\\{argv}[\|j]);{}$\6
${}\\{exit}({-}\T{3});{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\par
\U1.\fi

\N[62 commafree-eastman.w]{1}{3}The algorithm.
We think of $x$ as written cyclically, with $x_{n+j}=x_j$ for all~$j\ge0$.
The basic idea in the algorithm below is to also think of $x$ as partitioned
into $t\le n$ subwords by boundary markers $b_j$ where $0\le b_0<b_1<\cdots
<b_{t-1}<n$; then subword $y_j$ is $x_{b_j}x_{b_j+1}\ldots x_{b_{j+1}-1}$,
for $0\le j<t$, where $b_t=b_0$. If $t=1$, there's just one subword,
and it's a cyclic shift of~$x$. The number $t$ of subwords during each phase
will be odd.

Eastman's algorithm essentially begins with $b_j=j$ for $0\le j<n$, so that
$x$ is partitioned into $n$ subwords of length~1. It successively {\it removes%
\/}
boundary points until only one subword is left; that subword is the answer.
It operates in phases, so that all subwords during the $j$th phase have
length $3^{j-1}$ or more; thus at most $\lfloor\log_3n\rfloor$ phases
are needed. (For example, the case $n=9$ is ``dicey'' because it might
require two phases.)

The algorithm is based on comparison of adjacent subwords $y_{j-1}$ and~$y_j$.
If those subwords have the same length, we use lexicographic comparison;
otherwise we declare that the longer subword is bigger.

(After the first phase, all subwords not only have length $\ge3$, they also
always begin with a nonzero entry; in other words, $x_{b_j}>0$ for every
boundary marker~$b_j$. However, we won't need to use that fact explicitly.)

The algorithm can be described with terminology based on the topography
of Nevada: Say that $i$ is a {\it basin\/} if the subwords satisfy
$y_{i-1}>y_i\le y_{i+1}$. There must be at least one basin; otherwise all
the $y_j$ would be equal, and $x$ would equal one of its cyclic shifts.
We look at consecutive basins, $i$ and~$j$; this means that $i<j$ and that
$i$ and $j$ are basins, and that $i+1$ through $j-1$ are {\it not\/} basins.
If there's only one basin we have $j=i+t$.
The indices between consecutive basins are called {\it ranges\/}.

Since $t$ is odd, there's an odd number of consecutive basins for which
$j-i$ is odd. Each phase of Eastman's algorithm retains exactly one boundary
point in the range between such basins, and deletes all the others.
The retained point is the smallest $k=i+2l$ such that $y_k>y_{k+1}$.

(For example, suppose $i=2$ and $j=9$ are consecutive basins. Then we have
$y_1>y_2\le y_3\le\cdots\le y_q>y_{q+1}>\cdots>y_9\le y_{10}$, for some
range element $2<q<9$. We choose $k=4$ if $q=3$ or $q=4$,
$k=6$ if $q=5$ or $q=6$, and $k=8$ if $q=7$ or $q=8$.)

\Y\B\4\X3:Do Eastman's algorithm\X${}\E{}$\6
\X4:Initialize\X;\6
\&{for} ${}(\|p\K\T{1},\39\|t\K\|n;{}$ ${}\|t>\T{1};{}$ ${}\|t\K\\{tt},\39\|p%
\PP){}$\1\5
\X6:Do one phase of Eastman's algorithm, putting \PB{\\{tt}} boundary points
into \PB{\\{bb}}\X;\2\par
\U1.\fi

\M[111 commafree-eastman.w]{4}\B\X4:Initialize\X${}\E{}$\6
\&{for} ${}(\|j\K\|n;{}$ ${}\|j<\|n+\|n+\|n;{}$ ${}\|j\PP){}$\1\5
${}\|x[\|j]\K\|x[\|j-\|n];{}$\2\6
\&{for} ${}(\|j\K\T{0};{}$ ${}\|j<\|n+\|n+\|n;{}$ ${}\|j\PP){}$\1\5
${}\|b[\|j]\K\|j;{}$\2\6
${}\|t\K\|n{}$;\par
\U3.\fi

\M[116 commafree-eastman.w]{5}Here's a basic subroutine that returns 1 if
subword $y_{i-1}$ exceeds
subword~$y_i$, otherwise it returns~0.

\Y\B\4\X5:Subroutines\X${}\E{}$\6
\&{int} \\{compare}(\&{register} \&{int} \|i)\1\1\2\2\6
${}\{{}$\1\6
\&{register} \&{int} \|j;\7
\&{if} ${}(\|b[\|i]-\|b[\|i-\T{1}]\E\|b[\|i+\T{1}]-\|b[\|i]){}$\5
${}\{{}$\1\6
\&{for} ${}(\|j\K\T{0};{}$ ${}\|b[\|i]+\|j<\|b[\|i+\T{1}];{}$ ${}\|j\PP){}$\5
${}\{{}$\1\6
\&{if} ${}(\|x[\|b[\|i-\T{1}]+\|j]\E\|x[\|b[\|i]+\|j]){}$\1\5
\&{continue};\2\6
\&{return} ${}(\|x[\|b[\|i-\T{1}]+\|j]>\|x[\|b[\|i]+\|j]);{}$\6
\4${}\}{}$\2\6
\&{return} \T{0};\C{ $y_{i-1}=y_i$ }\6
\4${}\}{}$\2\6
\&{return} ${}(\|b[\|i]-\|b[\|i-\T{1}]>\|b[\|i+\T{1}]-\|b[\|i]);{}$\6
\4${}\}{}$\2\par
\U1.\fi

\M[132 commafree-eastman.w]{6}\B\X6:Do one phase of Eastman's algorithm,
putting \PB{\\{tt}} boundary points into \PB{\\{bb}}\X${}\E{}$\6
${}\{{}$\1\6
\&{for} ${}(\\{tt}\K\T{0},\39\|i\K\T{1};{}$ ${}\|i\Z\|t;{}$ ${}\|i\PP){}$\1\6
\&{if} (\\{compare}(\|i))\1\5
\&{break};\2\2\6
\&{if} ${}(\|i>\|t){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"The\ input\ is\ cyclic}\)\.{!\\n"});{}$\6
${}\\{exit}({-}\T{666});{}$\6
\4${}\}{}$\2\6
\&{for} ( ; ${}\\{compare}(\|i+\T{1});{}$ ${}\|i\PP){}$\1\5
;\C{ advance to the first basin }\2\6
\&{for} ( ; ${}\|i\Z\|t;{}$ ${}\|i\K\|j){}$\5
${}\{{}$\1\6
\&{for} ${}(\|q\K\|i+\T{1};{}$ ${}\\{compare}(\|q+\T{1})\E\T{0};{}$ ${}\|q%
\PP){}$\1\5
;\C{ climb the range }\2\6
\&{for} ${}(\|j\K\|q+\T{1};{}$ ${}\\{compare}(\|j+\T{1});{}$ ${}\|j\PP){}$\1\5
;\C{ advance to the next basin }\2\6
\&{if} ${}((\|j-\|i)\AND\T{1}){}$\1\5
\X7:Choose a boundary point to retain\X;\2\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"Phase\ \%d\ leaves"},\39\|p);{}$\6
\&{for} ${}(\|k\K\T{0};{}$ ${}\|k<\\{tt};{}$ ${}\|k\PP){}$\1\5
${}\|b[\|k]\K\\{bb}[\|k],\39\\{printf}(\.{"\ \%d"},\39\\{bb}[\|k]);{}$\2\6
\\{printf}(\.{"\\n"});\6
\&{for} ( ; ${}\|b[\|k-\\{tt}]<\|n+\|n;{}$ ${}\|k\PP){}$\1\5
${}\|b[\|k]\K\|b[\|k-\\{tt}]+\|n;{}$\2\6
\4${}\}{}$\2\par
\U3.\fi

\M[151 commafree-eastman.w]{7}\B\X7:Choose a boundary point to retain\X${}\E{}$%
\6
${}\{{}$\1\6
\&{if} ${}((\|q-\|i)\AND\T{1}){}$\1\5
${}\|q\PP;{}$\2\6
\&{if} ${}(\|q<\|t){}$\1\5
${}\\{bb}[\\{tt}\PP]\K\|b[\|q];{}$\2\6
\&{else}\5
${}\{{}$\1\6
\&{for} ${}(\|k\K\\{tt}\PP;{}$ ${}\|k>\T{0};{}$ ${}\|k\MM){}$\1\5
${}\\{bb}[\|k]\K\\{bb}[\|k-\T{1}];{}$\2\6
${}\\{bb}[\T{0}]\K\|b[\|q-\|t];{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\par
\U6.\fi

\N[161 commafree-eastman.w]{1}{8}Index.
\fi

\inx
\fin
\con
