\input cwebmac
\datethis




\N{1}{1}Hamiltonian cycles. This program finds all Hamiltonian cycles of
an undirected graph. [It's a slight revision of the program published
in my paper ``Mini-indexes for literate programs,'' {\sl Software---Concepts
and Tools\/ \bf15} (1994), 2--11.] The input graph should be in
Stanford GraphBase format, and should be named on the command line
as, for example, \.{foo.gb}. An optional second command-line parameter
is a modulus \PB{\|m}, which causes every $m$th solution to be printed.

We use a utility field to record the vertex degrees.

\Y\B\4\D$\\{deg}$ \5
$\|u.{}$\|I\par
\Y\B\8\#\&{include} \.{"gb\_graph.h"}\C{ the GraphBase data structures }\6
\8\#\&{include} \.{"gb\_save.h"}\C{ \PB{\\{restore\_graph}} }\6
\\{main}(\&{int} \\{argc}${},\39{}$\&{char} ${}{*}\\{argv}[\,]){}$\1\1\2\2\6
${}\{{}$\1\6
\&{Graph} ${}{*}\|g;{}$\6
\&{Vertex} ${}{*}\|x,{}$ ${}{*}\|y,{}$ ${}{*}\|z,{}$ ${}{*}\\{tmax};{}$\6
\&{register} \&{Vertex} ${}{*}\|t,{}$ ${}{*}\|u,{}$ ${}{*}\|v;{}$\6
\&{register} \&{Arc} ${}{*}\|a,{}$ ${}{*}\\{aa};{}$\6
\&{register} \&{int} \|d;\6
\&{Arc} ${}{*}\|b,{}$ ${}{*}\\{bb};{}$\6
\&{int} \\{count}${}\K\T{0};{}$\6
\&{int} \\{dmin}${},{}$ \\{modulus};\7
\X2:Process the command line, inputting the graph\X;\6
\X3:Prepare \PB{\|g} for backtracking, and find a vertex \PB{\|x} of minimum
degree\X;\6
\&{for} ${}(\|v\K\|g\MG\\{vertices};{}$ ${}\|v<\|g\MG\\{vertices}+\|g\MG\|n;{}$
${}\|v\PP){}$\1\5
${}\\{printf}(\.{"\ \%d"},\39\|v\MG\\{deg});{}$\2\6
\\{printf}(\.{"\\n"});\C{ TEMPORARY CHECK }\6
\&{if} ${}(\|x\MG\\{deg}<\T{2}){}$\5
${}\{{}$\1\6
${}\\{printf}(\.{"The\ minimum\ degree\ }\)\.{is\ \%d\ (vertex\ \%s)!\\n}\)%
\.{"},\39\|x\MG\\{deg},\39\|x\MG\\{name});{}$\6
\&{return} ${}{-}\T{1};{}$\6
\4${}\}{}$\2\6
\&{for} ${}(\|b\K\|x\MG\\{arcs};{}$ ${}\|b\MG\\{next};{}$ ${}\|b\K\|b\MG%
\\{next}){}$\1\6
\&{for} ${}(\\{bb}\K\|b\MG\\{next};{}$ \\{bb}; ${}\\{bb}\K\\{bb}\MG\\{next}){}$%
\5
${}\{{}$\1\6
${}\|v\K\|b\MG\\{tip};{}$\6
${}\|z\K\\{bb}\MG\\{tip};{}$\6
\X4:Find all simple paths of length \PB{$\|g\MG\|n-\T{2}$} from \PB{\|v} to %
\PB{\|z}, avoiding \PB{\|x}\X;\6
\4${}\}{}$\2\2\6
${}\\{printf}(\.{"Altogether\ \%d\ solut}\)\.{ions.\\n"},\39\\{count});{}$\6
\&{for} ${}(\|v\K\|g\MG\\{vertices};{}$ ${}\|v<\|g\MG\\{vertices}+\|g\MG\|n;{}$
${}\|v\PP){}$\1\5
${}\\{printf}(\.{"\ \%d"},\39\|v\MG\\{deg});{}$\2\6
\\{printf}(\.{"\\n"});\C{ TEMPORARY CHECK, SHOULD AGREE WITH FORMER VALUES }\6
\4${}\}{}$\2\par
\fi

\M{2}\B\X2:Process the command line, inputting the graph\X${}\E{}$\6
\&{if} ${}(\\{argc}>\T{1}){}$\1\5
${}\|g\K\\{restore\_graph}(\\{argv}[\T{1}]){}$;\5
\2\&{else}\1\5
${}\|g\K\NULL;{}$\2\6
\&{if} ${}(\\{argc}<\T{3}\V\\{sscanf}(\\{argv}[\T{2}],\39\.{"\%d"},\39{\AND}%
\\{modulus})\I\T{1}){}$\1\5
${}\\{modulus}\K\T{1000000000};{}$\2\6
\&{if} ${}(\R\|g\V\\{modulus}\Z\T{0}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Usage:\ \%s\ foo.gb\ [m}\)\.{odulus]\\n"},\39%
\\{argv}[\T{0}]);{}$\6
${}\\{exit}({-}\T{1});{}$\6
\4${}\}{}$\2\par
\U1.\fi

\M{3}Vertices that have already appeared in the path are ``taken,'' and
their \PB{\\{taken}} field is nonzero. Initially we make all those fields zero.

\Y\B\4\D$\\{taken}$ \5
$\|v.{}$\|I\par
\Y\B\4\X3:Prepare \PB{\|g} for backtracking, and find a vertex \PB{\|x} of
minimum degree\X${}\E{}$\6
$\\{dmin}\K\|g\MG\|n;{}$\6
\&{for} ${}(\|v\K\|g\MG\\{vertices};{}$ ${}\|v<\|g\MG\\{vertices}+\|g\MG\|n;{}$
${}\|v\PP){}$\5
${}\{{}$\1\6
${}\|v\MG\\{taken}\K\T{0};{}$\6
${}\|d\K\T{0};{}$\6
\&{for} ${}(\|a\K\|v\MG\\{arcs};{}$ \|a; ${}\|a\K\|a\MG\\{next}){}$\1\5
${}\|d\PP;{}$\2\6
${}\|v\MG\\{deg}\K\|d;{}$\6
\&{if} ${}(\|d<\\{dmin}){}$\1\5
${}\\{dmin}\K\|d,\39\|x\K\|v;{}$\2\6
\4${}\}{}$\2\par
\U1.\fi

\N{1}{4}The data structures. I use one simple rule to cut off unproductive
branches of the search tree: If one of the vertices we could move to next
is adjacent to only one other unused vertex, we must move to it now.

The moves will be recorded in the vertex array of \PB{\|g}. More precisely, the
\PB{\|k}th vertex of the path will be \PB{$\|t\MG\\{vert}$} when \PB{\|t} is
the \PB{\|k}th vertex of
the graph. If the move was not forced, \PB{$\|t\MG\\{ark}$} will point to the
Arc
record representing the edge from \PB{$\|t\MG\\{vert}$} to \PB{$(\|t+\T{1})\MG%
\\{vert}$}; otherwise
\PB{$\|t\MG\\{ark}$} will be \PB{$\NULL$}.

This program is a typical backtrack program. I am more comfortable doing
it with labels and goto statements than with while loops, but some day
I may learn my lesson.

\Y\B\4\D$\\{vert}$ \5
$\|w.{}$\|V\par
\B\4\D$\\{ark}$ \5
$\|x.{}$\|A\par
\Y\B\4\X4:Find all simple paths of length \PB{$\|g\MG\|n-\T{2}$} from \PB{\|v}
to \PB{\|z}, avoiding \PB{\|x}\X${}\E{}$\6
$\|t\K\|g\MG\\{vertices}{}$;\5
${}\\{tmax}\K\|t+\|g\MG\|n-\T{1};{}$\6
${}\|x\MG\\{taken}\K\T{1}{}$;\5
${}\|t\MG\\{vert}\K\|x;{}$\6
${}\|t\MG\\{ark}\K\NULL;{}$\6
\4\\{advance}:\5
\X5:Increase \PB{\|t} and update the data structures to show that vertex \PB{%
\|v} is now taken; \PB{\&{goto} \\{backtrack}} if no further moves are possible%
\X;\6
\4\\{try}:\5
\X7:Look at edge \PB{\|a} and its successors, advancing if it is a valid move%
\X;\6
\4\\{restore}:\5
\X6:Downdate the data structures to the state they were in when level \PB{\|t}
was entered\X;\6
\4\\{backtrack}:\5
\X8:Decrease \PB{\|t}, if possible, and try the next possibility; or \PB{%
\&{goto} \\{done}}\X;\6
\\{done}:\par
\U1.\fi

\M{5}\B\X5:Increase \PB{\|t} and update the data structures to show that vertex
\PB{\|v} is now taken; \PB{\&{goto} \\{backtrack}} if no further moves are
possible\X${}\E{}$\6
$\|t\PP;{}$\6
${}\|t\MG\\{vert}\K\|v;{}$\6
${}\|v\MG\\{taken}\K\T{1};{}$\6
\&{if} ${}(\|v\E\|z){}$\5
${}\{{}$\1\6
\&{if} ${}(\|t\E\\{tmax}){}$\1\5
\X9:Record a solution\X;\2\6
\&{goto} \\{backtrack};\6
\4${}\}{}$\2\6
\&{for} ${}(\\{aa}\K\|v\MG\\{arcs},\39\|y\K\NULL;{}$ \\{aa}; ${}\\{aa}\K\\{aa}%
\MG\\{next}){}$\5
${}\{{}$\1\6
${}\|u\K\\{aa}\MG\\{tip};{}$\6
${}\|d\K\|u\MG\\{deg}-\T{1};{}$\6
\&{if} ${}(\|d\E\T{1}\W\|u\MG\\{taken}\E\T{0}){}$\5
${}\{{}$\1\6
\&{if} (\|y)\1\5
\&{goto} \\{restore};\C{ restoration will stop at \PB{\\{aa}} }\2\6
${}\|y\K\|u;{}$\6
\4${}\}{}$\2\6
${}\|u\MG\\{deg}\K\|d;{}$\6
\4${}\}{}$\2\6
\&{if} (\|y)\5
${}\{{}$\1\6
${}\|t\MG\\{ark}\K\NULL;{}$\6
${}\|v\K\|y;{}$\6
\&{goto} \\{advance};\6
\4${}\}{}$\2\6
${}\|a\K\|v\MG\\{arcs}{}$;\par
\U4.\fi

\M{6}\B\X6:Downdate the data structures to the state they were in when level %
\PB{\|t} was entered\X${}\E{}$\6
\&{for} ${}(\|a\K\|t\MG\\{vert}\MG\\{arcs};{}$ ${}\|a\I\\{aa};{}$ ${}\|a\K\|a%
\MG\\{next}){}$\1\5
${}\|a\MG\\{tip}\MG\\{deg}\PP{}$;\2\par
\U4.\fi

\M{7}\B\X7:Look at edge \PB{\|a} and its successors, advancing if it is a valid
move\X${}\E{}$\6
\&{while} (\|a)\5
${}\{{}$\1\6
${}\|v\K\|a\MG\\{tip};{}$\6
\&{if} ${}(\|v\MG\\{taken}\E\T{0}){}$\5
${}\{{}$\1\6
${}\|t\MG\\{ark}\K\|a;{}$\6
\&{goto} \\{advance};\6
\4${}\}{}$\2\6
${}\|a\K\|a\MG\\{next};{}$\6
\4${}\}{}$\2\6
\4\\{restore\_all}:\5
${}\\{aa}\K\NULL{}$;\C{ all moves tried; we fall through to \PB{\\{restore}} }%
\par
\U4.\fi

\M{8}\B\X8:Decrease \PB{\|t}, if possible, and try the next possibility; or %
\PB{\&{goto} \\{done}}\X${}\E{}$\6
$\|t\MG\\{vert}\MG\\{taken}\K\T{0};{}$\6
${}\|t\MM;{}$\6
\&{if} ${}(\|t\MG\\{ark}){}$\5
${}\{{}$\1\6
${}\|a\K\|t\MG\\{ark}\MG\\{next};{}$\6
\&{goto} \\{try};\6
\4${}\}{}$\2\6
\&{if} ${}(\|t\E\|g\MG\\{vertices}){}$\1\5
\&{goto} \\{done};\2\6
\&{goto} \\{restore\_all};\C{ the move was forced }\par
\U4.\fi

\M{9}\B\X9:Record a solution\X${}\E{}$\6
${}\{{}$\1\6
${}\\{count}\PP;{}$\6
\&{if} ${}(\\{count}\MOD\\{modulus}\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{printf}(\.{"\%d:\ "},\39\\{count});{}$\6
\&{for} ${}(\|u\K\|g\MG\\{vertices};{}$ ${}\|u\Z\\{tmax};{}$ ${}\|u\PP){}$\1\5
${}\\{printf}(\.{"\%s\ "},\39\|u\MG\\{vert}\MG\\{name});{}$\2\6
\\{printf}(\.{"\\n"});\6
\4${}\}{}$\2\6
\4${}\}{}$\2\par
\U5.\fi

\N{1}{10}Index.
\fi

\inx
\fin
\con
