\input cwebmac
\hypertextrue\srcloctrue
\datethis




\N[4 sham.w]{1}{1}Symmetric Hamiltonian cycles. This program finds all
Hamiltonian cycles of
an undirected graph in which the mapping $v\mapsto N-1-v$ is an
automorphism, such that the same automorphism also applies to the cycle.

We use a utility field to record the vertex degrees.

\Y\B\4\D$\\{deg}$ \5
$\|u.{}$\|I\par
\B\4\D$\\{mm}$ \5
\T{8}\C{ should be even }\par
\B\4\D$\\{nn}$ \5
\T{9}\par
\Y\B\8\#\&{include} \.{"gb\_graph.h"}\C{ the GraphBase data structures }\6
\8\#\&{include} \.{"gb\_basic.h"}\C{ standard graphs }\6
\\{main}(\,)\1\1\2\2\6
${}\{{}$\1\6
\&{Graph} ${}{*}\|g\K\\{board}(\\{mm},\39\\{nn},\39\T{0},\39\T{0},\39\T{5},\39%
\T{0},\39\T{0}){}$;\C{ knight moves on rectangular chessboard }\6
\&{Vertex} ${}{*}\|x,{}$ ${}{*}\|z,{}$ ${}{*}\\{tmax};{}$\6
\&{register} \&{Vertex} ${}{*}\|t,{}$ ${}{*}\|u,{}$ ${}{*}\|v;{}$\6
\&{register} \&{Arc} ${}{*}\|a,{}$ ${}{*}\\{aa},{}$ ${}{*}\\{yy};{}$\6
\&{register} \&{int} \|d;\6
\&{Arc} ${}{*}\|b,{}$ ${}{*}\\{bb};{}$\6
\&{int} \\{count}${}\K\T{0},{}$ \\{dcount}${}\K\T{0};{}$\6
\&{int} \\{dmin};\7
\X2:Reduce \PB{\|g} to half size\X;\6
\X4:Prepare \PB{\|g} for backtracking, and find a vertex \PB{\|x} of minimum
degree\X;\6
\&{for} ${}(\|v\K\|g\MG\\{vertices};{}$ ${}\|v<\|g\MG\\{vertices}+\|g\MG\|n;{}$
${}\|v\PP){}$\1\5
${}\\{printf}(\.{"\ \%d"},\39\|v\MG\\{deg});{}$\2\6
\\{printf}(\.{"\\n"});\C{ TEMPORARY CHECK }\6
\&{if} ${}(\|x\MG\\{deg}<\T{2}){}$\5
${}\{{}$\1\6
${}\\{printf}(\.{"The\ minimum\ degree\ }\)\.{is\ \%d\ (vertex\ \%s)!\\n}\)%
\.{"},\39\|x\MG\\{deg},\39\|x\MG\\{name});{}$\6
\&{return} ${}{-}\T{1};{}$\6
\4${}\}{}$\2\6
\&{for} ${}(\|b\K\|x\MG\\{arcs};{}$ ${}\|b\MG\\{next};{}$ ${}\|b\K\|b\MG%
\\{next}){}$\1\6
\&{for} ${}(\\{bb}\K\|b\MG\\{next};{}$ \\{bb}; ${}\\{bb}\K\\{bb}\MG\\{next}){}$%
\5
${}\{{}$\1\6
${}\|a\K\|b;{}$\6
${}\|z\K\\{bb}\MG\\{tip};{}$\6
\X5:Find all simple paths of length \PB{$\|g\MG\|n-\T{2}$} from \PB{$\|a\MG%
\\{tip}$} to \PB{\|z}, avoiding \PB{\|x}\X;\6
\4${}\}{}$\2\2\6
${}\\{printf}(\.{"Altogether\ \%d\ solut}\)\.{ions\ and\ \%d\ wannabee}\)\.{s.%
\\n"},\39\\{count},\39\\{dcount});{}$\6
\&{for} ${}(\|v\K\|g\MG\\{vertices};{}$ ${}\|v<\|g\MG\\{vertices}+\|g\MG\|n;{}$
${}\|v\PP){}$\1\5
${}\\{printf}(\.{"\ \%d"},\39\|v\MG\\{deg});{}$\2\6
\\{printf}(\.{"\\n"});\C{ TEMPORARY CHECK, SHOULD AGREE WITH FORMER VALUES }\6
\4${}\}{}$\2\par
\fi

\M[47 sham.w]{2}We identify each vertex with its symmetric mate, and set the
length
of an arc to 1 if the arc crosses to the mate instead of staying in the
same class.

Multiple arcs and self-loops can be introduced in this step.

\Y\B\4\D$\\{mate}(\|v)$ \5
(\&{Vertex} ${}{*})({}$((\&{unsigned} \&{long}) \|g${}\MG\\{vertices})+{}$((%
\&{unsigned} \&{long})(${}\|g\MG\\{vertices}+\|g\MG\|n-\T{1}))-{}$((%
\&{unsigned} \&{long}) \|v))\par
\Y\B\4\X2:Reduce \PB{\|g} to half size\X${}\E{}$\6
\&{for} ${}(\|v\K\|g\MG\\{vertices};{}$ ${}\\{mate}(\|v)>\|v;{}$ ${}\|v\PP){}$%
\1\6
\&{for} ${}(\|a\K\|v\MG\\{arcs};{}$ \|a; ${}\|a\K\|a\MG\\{next}){}$\5
${}\{{}$\1\6
${}\|u\K\\{mate}(\|a\MG\\{tip});{}$\6
\&{if} ${}(\|u>\|a\MG\\{tip}){}$\1\5
${}\|a\MG\\{len}\K\T{0};{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\|a\MG\\{len}\K\T{1};{}$\6
${}\|a\MG\\{tip}\K\|u;{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\2\6
${}\|g\MG\|n\MRL{{/}{\K}}\T{2}{}$;\par
\U1.\fi

\M[69 sham.w]{3}Self-loops caused a subtle bug (my test for \PB{$\|v\MG\\{deg}%
\E\T{1}$} below
failed), and they are of no interest in Hamiltonian circuits. So
I'm now getting rid of them.

\Y\B\4\X3:Remove self-loops\X${}\E{}$\6
\&{for} ${}(\|v\K\|g\MG\\{vertices};{}$ ${}\|v<\|g\MG\\{vertices}+\|g\MG\|n;{}$
${}\|v\PP){}$\1\6
\&{for} ${}(\|a\K\|v\MG\\{arcs},\39\\{aa}\K\NULL;{}$ \|a; ${}\|a\K\|a\MG%
\\{next}){}$\1\6
\&{if} ${}(\|a\MG\\{tip}\E\|v){}$\5
${}\{{}$\1\6
\&{if} (\\{aa})\1\5
${}\\{aa}\MG\\{next}\K\|a\MG\\{next};{}$\2\6
\&{else}\1\5
${}\|v\MG\\{arcs}\K\|a\MG\\{next};{}$\2\6
\4${}\}{}$\2\6
\&{else}\1\5
${}\\{aa}\K\|a{}$;\2\2\2\par
\U4.\fi

\M[81 sham.w]{4}Vertices that have already appeared in the path are ``taken,''
and
their \PB{\\{taken}} field is nonzero. Initially we make all those fields zero.

\Y\B\4\D$\\{taken}$ \5
$\|v.{}$\|I\par
\Y\B\4\X4:Prepare \PB{\|g} for backtracking, and find a vertex \PB{\|x} of
minimum degree\X${}\E{}$\6
\X3:Remove self-loops\X;\6
${}\\{dmin}\K\|g\MG\|n;{}$\6
\&{for} ${}(\|v\K\|g\MG\\{vertices};{}$ ${}\|v<\|g\MG\\{vertices}+\|g\MG\|n;{}$
${}\|v\PP){}$\5
${}\{{}$\1\6
${}\|v\MG\\{taken}\K\T{0};{}$\6
${}\|d\K\T{0};{}$\6
\&{for} ${}(\|a\K\|v\MG\\{arcs};{}$ \|a; ${}\|a\K\|a\MG\\{next}){}$\1\5
${}\|d\PP;{}$\2\6
${}\|v\MG\\{deg}\K\|d;{}$\6
\&{if} ${}(\|d<\\{dmin}){}$\1\5
${}\\{dmin}\K\|d,\39\|x\K\|v;{}$\2\6
\4${}\}{}$\2\par
\U1.\fi

\N[97 sham.w]{1}{5}The data structures. I use one simple rule to cut off
unproductive
branches of the search tree: If one of the vertices we could move to next
is adjacent to only one other unused vertex, we must move to it now.

The moves will be recorded in the vertex array of \PB{\|g}. More precisely, the
\PB{\|k}th arc of the path will be \PB{$\|t\MG\\{ark}$} when \PB{\|t} is the %
\PB{\|k}th vertex of
the graph.

This program is a typical backtrack program. I am more comfortable doing
it with labels and goto statements than with while loops, but some day
I may learn my lesson.

\Y\B\4\D$\\{ark}$ \5
$\|x.{}$\|A\par
\Y\B\4\X5:Find all simple paths of length \PB{$\|g\MG\|n-\T{2}$} from \PB{$\|a%
\MG\\{tip}$} to \PB{\|z}, avoiding \PB{\|x}\X${}\E{}$\6
$\|v\K\|a\MG\\{tip};{}$\6
${}\|t\K\|g\MG\\{vertices}{}$;\5
${}\\{tmax}\K\|t+\|g\MG\|n-\T{1};{}$\6
${}\|x\MG\\{taken}\K\T{1};{}$\6
${}\|a\MG\\{len}\MRL{+{\K}}\T{4}{}$;\C{ the first move is ``forced'' }\6
\4\\{advance}:\5
\X6:Increase \PB{\|t} and update the data structures to show that vertex \PB{%
\|v} is now taken; \PB{\&{goto} \\{backtrack}} if no further moves are possible%
\X;\6
\&{try} : \X8:Look at edge \PB{\|a} and its successors, advancing if it is a
valid move\X;\6
\4\\{restore}:\5
\X7:Downdate the data structures to the state they were in when level \PB{\|t}
was entered\X;\6
\4\\{backtrack}:\5
\X9:Decrease \PB{\|t}, if possible, and try the next possibility; or \PB{%
\&{goto} \\{done}}\X;\6
\\{done}:\par
\U1.\fi

\M[126 sham.w]{6}\B\X6:Increase \PB{\|t} and update the data structures to show
that vertex \PB{\|v} is now taken; \PB{\&{goto} \\{backtrack}} if no further
moves are possible\X${}\E{}$\6
$\|t\MG\\{ark}\K\|a;{}$\6
${}\|t\PP;{}$\6
${}\|v\K\|a\MG\\{tip};{}$\6
${}\|v\MG\\{taken}\K\T{1};{}$\6
\&{if} ${}(\|v\E\|z){}$\5
${}\{{}$\1\6
\&{if} ${}(\|t\E\\{tmax}\W\|v\MG\\{deg}\E\T{1}){}$\1\5
\X10:Record a solution\X;\2\6
\&{goto} \\{backtrack};\6
\4${}\}{}$\2\6
${}\\{yy}\K\NULL{}$;\C{ \PB{\\{yy}} is a forced arc, if any exist }\6
\&{for} ${}(\\{aa}\K\|v\MG\\{arcs};{}$ \\{aa}; ${}\\{aa}\K\\{aa}\MG\\{next}){}$%
\5
${}\{{}$\1\6
${}\|u\K\\{aa}\MG\\{tip};{}$\6
${}\|d\K\|u\MG\\{deg}-\T{1};{}$\6
\&{if} ${}(\|d\E\T{1}\W\|u\MG\\{taken}\E\T{0}){}$\5
${}\{{}$\1\6
\&{if} (\\{yy})\1\5
\&{goto} \\{restore};\C{ restoration will stop at \PB{\\{aa}} }\2\6
${}\\{yy}\K\\{aa};{}$\6
\4${}\}{}$\2\6
${}\|u\MG\\{deg}\K\|d;{}$\6
\4${}\}{}$\2\6
\&{if} (\\{yy})\5
${}\{{}$\1\6
${}\|a\K\\{yy};{}$\6
${}\|a\MG\\{len}\MRL{+{\K}}\T{4};{}$\6
\&{goto} \\{advance};\6
\4${}\}{}$\2\6
${}\|a\K\|v\MG\\{arcs}{}$;\par
\U5.\fi

\M[152 sham.w]{7}\B\X7:Downdate the data structures to the state they were in
when level \PB{\|t} was entered\X${}\E{}$\6
\&{for} ${}(\|a\K(\|t-\T{1})\MG\\{ark}\MG\\{tip}\MG\\{arcs};{}$ ${}\|a\I%
\\{aa};{}$ ${}\|a\K\|a\MG\\{next}){}$\1\5
${}\|a\MG\\{tip}\MG\\{deg}\PP{}$;\2\par
\U5.\fi

\M[156 sham.w]{8}\B\X8:Look at edge \PB{\|a} and its successors, advancing if
it is a valid move\X${}\E{}$\6
\&{while} (\|a)\5
${}\{{}$\1\6
\&{if} ${}(\|a\MG\\{tip}\MG\\{taken}\E\T{0}){}$\5
${}\{{}$\1\6
${}\|a\MG\\{len}\MRL{+{\K}}\T{2}{}$;\C{ oops, this is unnecessary residue of {%
\sc SHAMR} }\6
\&{goto} \\{advance};\6
\4${}\}{}$\2\6
${}\|a\K\|a\MG\\{next};{}$\6
\4${}\}{}$\2\6
\4\\{restore\_all}:\5
${}\\{aa}\K\NULL{}$;\C{ all moves tried; we fall through to \PB{\\{restore}} }%
\par
\U5.\fi

\M[166 sham.w]{9}Here we come to a subtle point. When a forced move is a
duplicated arc,
we want to continue with the duplicate arc; we don't want to backtrack!

But that isn't the most subtle part. It turns out that we want to
consider the duplicate arc {\it previous\/} to the one that worked.
(That one should really have been considered forced; if on the other
hand the first of two duplicate arcs is selected, the second one will
decrease the degree to zero and cannot lead to a complete tour, so
we don't want to reconsider it.) Get it? The present logic works
only when there are at most two duplicate arcs.

\Y\B\4\X9:Decrease \PB{\|t}, if possible, and try the next possibility; or \PB{%
\&{goto} \\{done}}\X${}\E{}$\6
$\|t\MM;{}$\6
${}\|a\K\|t\MG\\{ark};{}$\6
${}\|a\MG\\{tip}\MG\\{taken}\K\T{0};{}$\6
${}\|d\K\|a\MG\\{len};{}$\6
${}\|a\MG\\{len}\MRL{\AND{\K}}\T{1};$ \&{if} ${}(\|d<\T{4})$ $\{$ $\|a\K\|a\MG%
\\{next};$ \&{goto} \6
\&{try}\1\5
;\2\6
$\}$ \6
\&{if} ${}(\|t\E\|g\MG\\{vertices}){}$\1\5
\&{goto} \\{done};\2\6
\&{for} ${}(\\{aa}\K(\|t-\T{1})\MG\\{ark}\MG\\{tip}\MG\\{arcs};{}$ ${}\\{aa}\I%
\|a;{}$ ${}\\{aa}\K\\{aa}\MG\\{next}){}$\1\6
\&{if} ${}(\\{aa}\MG\\{tip}\E\|a\MG\\{tip}){}$\5
${}\{{}$\1\6
${}\\{aa}\MG\\{len}\MRL{+{\K}}\T{4};{}$\6
${}\|a\K\\{aa};{}$\6
\&{goto} \\{advance};\6
\4${}\}{}$\2\2\6
\&{goto} \\{restore\_all};\C{ the move was forced }\par
\U5.\fi

\M[195 sham.w]{10}\B\X10:Record a solution\X${}\E{}$\6
${}\{{}$\1\6
\&{int} \|s${}\K\T{0};{}$\7
\&{for} ${}(\|u\K\|g\MG\\{vertices};{}$ ${}\|u<\\{tmax};{}$ ${}\|u\PP){}$\1\5
${}\|s\MRL{{\XOR}{\K}}\|u\MG\\{ark}\MG\\{len}\AND\T{1};{}$\2\6
\&{if} (\|s)\5
${}\{{}$\1\6
${}\\{count}\PP;{}$\6
\&{if} ${}(\\{count}\MOD\T{100000}\E\T{0}){}$\5
${}\{{}$\C{ use 100000 for the $8\times8$ }\1\6
${}\\{printf}(\.{"\%d:\ \%s"},\39\\{count},\39\|x\MG\\{name});{}$\6
\&{for} ${}(\|u\K\|g\MG\\{vertices};{}$ ${}\|u<\\{tmax};{}$ ${}\|u\PP){}$\1\5
${}\\{printf}(\.{"\%s\%s"},\39\|u\MG\\{ark}\MG\\{len}\AND\T{1}\?\.{"*"}:\.{"\
"},\39\|u\MG\\{ark}\MG\\{tip}\MG\\{name});{}$\2\6
\\{printf}(\.{"\\n"});\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{dcount}\PP;{}$\6
\&{if} ${}(\\{dcount}\MOD\T{100000}\E\T{0}){}$\5
${}\{{}$\C{ use 1 for small cases }\1\6
${}\\{printf}(\.{">\%d:\ \%s"},\39\\{dcount},\39\|x\MG\\{name});{}$\6
\&{for} ${}(\|u\K\|g\MG\\{vertices};{}$ ${}\|u<\\{tmax};{}$ ${}\|u\PP){}$\1\5
${}\\{printf}(\.{"\%s\%s"},\39\|u\MG\\{ark}\MG\\{len}\AND\T{1}\?\.{"*"}:\.{"\
"},\39\|u\MG\\{ark}\MG\\{tip}\MG\\{name});{}$\2\6
\\{printf}(\.{"\\n"});\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\par
\U6.\fi

\N[217 sham.w]{1}{11}Index.
\fi

\inx
\fin
\con
