\input cwebmac
\srcloctrue
\datethis

\N[2 xgdance.w]{1}{1}Generalized exact cover.
This program implements an extension of
the algorithm discussed in my paper about ``dancing
links.'' I hacked it together from the {\mc XCOVER} program that I wrote in
1994; I apologize for not having time to apply spit and polish.

Given a matrix whose elements are 0 or 1, the problem in that paper was to
find all subsets of its rows whose sum is at most~1 in all columns and
{\it exactly\/}~1 in all ``primary'' columns. The matrix is specified
in the standard input file as follows: Each column has a symbolic name,
up to seven characters long. The first line of input contains
the names of all primary columns, followed by `\.{\char"7C}', followed by
the names of all other columns.
(If all columns are primary, the~`\.{\char"7C}' may be omitted.)
The remaining lines represent the rows, by listing the columns where 1 appears.

Here I extend the idea so that nonprimary columns have a different sort
of restriction: If a row specifies a ``color'' in a nonprimary column,
it rules out rows of all other colors, but any number of rows with the
same color are allowed. (The previous situation was the special case
in which all rows had a different color.)
If \.{xx} is a column name, a specification like \.{xx:a} as part of a row
stands for color~\.a in column~\.{xx}. Each color is specified by a
single character.

Also---very important---I assume here that the input data is totally
symmetric in all colors (except perhaps the smallest). In other words,
whenever there's a row specifying color attributes $(\chi_1,\ldots,\chi_k)$
in columns $(c_1,\ldots,c_k)$, there's also a row that specifies
attributes $(\pi\chi_1,\ldots,\pi\chi_k)$ for those colors, for every
permutation $\pi$ that fixes the smallest color. Then if there are
$t$~colors, there will be $(t-1)!$ equivalent solutions; this program
speeds things up by finding exactly one of them.
The smallest color is assumed to be \PB{\.{'a'}},
and the others are assumed to be \PB{\.{'b'}}, \PB{\.{'c'}}, etc.
This program is identical to {\mc GDANCE} except for places where
it refers to the new variables \PB{\\{color\_thresh}}, \PB{\\{conflict}}, and %
\PB{\\{cthresh}}.

The program prints the number of solutions and the total number of link
updates. It also prints every $n$th solution, if the integer command
line argument $n$ is given. A second command-line argument causes the
full search tree to be printed, and a third argument makes the output
even more verbose.

\Y\B\4\D$\\{max\_level}$ \5
\T{1000}\C{ at most this many rows in a solution }\par
\B\4\D$\\{max\_degree}$ \5
\T{1000}\C{ at most this many branches per search tree node }\par
\B\4\D$\\{max\_cols}$ \5
\T{1000}\C{ at most this many columns }\par
\B\4\D$\\{max\_nodes}$ \5
\T{200000}\C{ at most this many nonzero elements in the matrix }\par
\Y\B\8\#\&{include} \.{<stdio.h>}\6
\8\#\&{include} \.{<ctype.h>}\6
\8\#\&{include} \.{<string.h>}\6
\X3:Type definitions\X\6
\X2:Global variables\X\6
\X6:Subroutines\X;\7
${}\\{main}(\\{argc},\39\\{argv}){}$\1\1\6
\&{int} \\{argc};\6
\&{char} ${}{*}\\{argv}[\,];\2\2{}$\6
${}\{{}$\1\6
\X10:Local variables\X;\6
${}\\{verbose}\K\\{argc}-\T{1};{}$\6
\&{if} (\\{verbose})\1\5
${}\\{sscanf}(\\{argv}[\T{1}],\39\.{"\%d"},\39{\AND}\\{spacing});{}$\2\6
\X7:Initialize the data structures\X;\6
\X12:Backtrack through all solutions\X;\6
${}\\{printf}(\.{"Altogether\ \%d\ solut}\)\.{ions,\ after\ \%u\ updat}\)\.{es\
and\ \%u\ cleansings}\)\.{.\\n"},\39\\{count},\39\\{updates},\39\\{purifs});{}$%
\6
\&{if} (\\{verbose})\1\5
\X23:Print a profile of the search tree\X;\2\6
\\{exit}(\T{0});\6
\4${}\}{}$\2\par
\fi

\M[74 xgdance.w]{2}\B\X2:Global variables\X${}\E{}$\6
\&{int} \\{verbose};\C{ $>0$ to show solutions, $>1$ to show partial ones too }%
\6
\&{int} \\{count}${}\K\T{0}{}$;\C{ number of solutions found so far }\6
\&{unsigned} \&{int} \\{updates};\C{ number of times we deleted a list element
}\6
\&{unsigned} \&{int} \\{purifs};\C{ number of times we purified a list element
}\6
\&{int} \\{spacing}${}\K\T{1}{}$;\C{ if \PB{\\{verbose}}, we output solutions
when \PB{$\\{count}\MOD\\{spacing}\E\T{0}$} }\6
\&{int} \\{profile}[\\{max\_level}][\\{max\_degree}];\C{ tree nodes of given
level and degree }\6
\&{unsigned} \&{int} \\{upd\_prof}[\\{max\_level}];\C{ updates at a given level
}\6
\&{unsigned} \&{int} \\{pur\_prof}[\\{max\_level}];\C{ purifications at a given
level }\6
\&{int} \\{maxb}${}\K\T{0}{}$;\C{ maximum branching factor actually needed }\6
\&{int} \\{maxl}${}\K\T{0}{}$;\C{ maximum level actually reached }\par
\As8\ET14.
\U1.\fi

\N[86 xgdance.w]{1}{3}Data structures.
Each column of the input matrix is represented by a \&{column} struct,
and each row is represented as a linked list of \&{node} structs. There's one
node for each nonzero entry in the matrix.

More precisely, the nodes are linked circularly within each row, in
both directions. The nodes are also linked circularly within each column;
the column lists each include a header node, but the row lists do not.
Column header nodes are part of a \&{column} struct, which
contains further info about the column.

Each node contains five fields. Four are the pointers of doubly linked lists,
already mentioned; the fifth points to the column containing the node.

\Y\B\4\X3:Type definitions\X${}\E{}$\6
\&{typedef} \&{struct} \&{node\_struct} ${}\{{}$\1\6
\&{struct} \&{node\_struct} ${}{*}\\{left},{}$ ${}{*}\\{right}{}$;\C{
predecessor and successor in row }\6
\&{struct} \&{node\_struct} ${}{*}\\{up},{}$ ${}{*}\\{down}{}$;\C{ predecessor
and successor in column }\6
\&{struct} \&{col\_struct} ${}{*}\\{col}{}$;\C{ the column containing this node
}\6
\&{int} \\{color};\C{ color, if specified }\2\6
${}\}{}$ \&{node};\par
\A4.
\U1.\fi

\M[110 xgdance.w]{4}Each \&{column} struct contains five fields:
The \PB{\\{head}} is a node that stands at the head of its list of nodes;
the \PB{\\{len}} tells the length of that list of nodes, not counting the
header;
the \PB{\\{name}} is a one-, two-, or \dots\ or seven-letter identifier;
\PB{\\{next}} and \PB{\\{prev}} point to adjacent columns, when this
column is part of a doubly linked list.

As backtracking proceeds, nodes
will be deleted from column lists when their row has been blocked by
other rows in the partial solution.
But when backtracking is complete, the data structures will be
restored to their original state.

\Y\B\4\X3:Type definitions\X${}\mathrel+\E{}$\6
\&{typedef} \&{struct} \&{col\_struct} ${}\{{}$\1\6
\&{node} \\{head};\C{ the list header }\6
\&{int} \\{len};\C{ the number of non-header items currently in this column's
list }\6
\&{char} \\{name}[\T{8}];\C{ symbolic identification of the column, for
printing }\6
\&{struct} \&{col\_struct} ${}{*}\\{prev},{}$ ${}{*}\\{next}{}$;\C{ neighbors
of this column }\6
\&{int} \\{color\_thresh};\C{ used for backing up }\2\6
${}\}{}$ \&{column};\par
\fi

\M[132 xgdance.w]{5}One \PB{\&{column}} struct is called the root. It serves as
the head of the
list of columns that need to be covered, and is identifiable by the fact
that its \PB{\\{name}} is empty.

\Y\B\4\D$\\{root}$ \5
\\{col\_array}[\T{0}]\C{ gateway to the unsettled columns }\par
\fi

\M[138 xgdance.w]{6}A row is identified not by name but by the names of the
columns it contains.
Here is a routine that prints a row, given a pointer to any of its
nodes. It also prints the position of the given node in its column.

\Y\B\4\X6:Subroutines\X${}\E{}$\6
\\{print\_row}(\|p)\1\1\6
\&{node} ${}{*}\|p;\2\2{}$\6
${}\{{}$\5
\1\&{register} \&{node} ${}{*}\|q\K\|p;{}$\6
\&{register} \&{int} \|k;\7
\&{do}\5
${}\{{}$\1\6
${}\\{printf}(\.{"\ \%s"},\39\|q\MG\\{col}\MG\\{name});{}$\6
\&{if} ${}(\|q\MG\\{color}){}$\1\5
\X27:Print the color of node \PB{\|q}\X;\2\6
${}\|q\K\|q\MG\\{right};{}$\6
\4${}\}{}$\2\5
\&{while} ${}(\|q\I\|p);{}$\6
\&{for} ${}(\|q\K\|p\MG\\{col}\MG\\{head}.\\{down},\39\|k\K\T{1};{}$ ${}\|q\I%
\|p;{}$ ${}\|k\PP){}$\1\6
\&{if} ${}(\|q\E{\AND}(\|p\MG\\{col}\MG\\{head})){}$\5
${}\{{}$\1\6
\\{printf}(\.{"\\n"});\5
\&{return};\C{ row not in its column! }\6
\4${}\}{}$\5
\2\&{else}\1\5
${}\|q\K\|q\MG\\{down};{}$\2\2\6
${}\\{printf}(\.{"\ (\%d\ of\ \%d)\\n"},\39\|k,\39\|p\MG\\{col}\MG\\{len});{}$\6
\4${}\}{}$\2\par
\As15, 16, 25, 26\ETs28.
\U1.\fi

\N[159 xgdance.w]{1}{7}Inputting the matrix.
Brute force is the rule in this part of the program.

\Y\B\4\X7:Initialize the data structures\X${}\E{}$\6
\X9:Read the column names\X;\6
\X11:Read the rows\X;\par
\U1.\fi

\M[166 xgdance.w]{8}\B\D$\\{buf\_size}$ \5
$\T{4}*\\{max\_cols}+{}$\T{3}\C{ upper bound on input line length }\par
\Y\B\4\X2:Global variables\X${}\mathrel+\E{}$\6
\&{column} ${}\\{col\_array}[\\{max\_cols}+\T{2}]{}$;\C{ place for column
records }\6
\&{node} \\{node\_array}[\\{max\_nodes}];\C{ place for nodes }\6
\&{char} \\{buf}[\\{buf\_size}];\par
\fi

\M[173 xgdance.w]{9}\B\D$\\{panic}(\|m)$ \6
${}\{{}$\5
\1${}\\{fprintf}(\\{stderr},\39\.{"\%s!\\n\%s"},\39\|m,\39\\{buf}){}$;\5
${}\\{exit}({-}\T{1}){}$;\5
${}\}{}$\2\par
\Y\B\4\X9:Read the column names\X${}\E{}$\6
$\\{cur\_col}\K\\{col\_array}+\T{1};{}$\6
${}\\{fgets}(\\{buf},\39\\{buf\_size},\39\\{stdin});{}$\6
\&{if} ${}(\\{buf}[\\{strlen}(\\{buf})-\T{1}]\I\.{'\\n'}){}$\1\5
\\{panic}(\.{"Input\ line\ too\ long}\)\.{"});\2\6
\&{for} ${}(\|p\K\\{buf},\39\\{primary}\K\T{1};{}$ ${}{*}\|p;{}$ ${}\|p\PP){}$\5
${}\{{}$\1\6
\&{while} ${}(\\{isspace}({*}\|p)){}$\1\5
${}\|p\PP;{}$\2\6
\&{if} ${}(\R{*}\|p){}$\1\5
\&{break};\2\6
\&{if} ${}({*}\|p\E\.{'|'}){}$\5
${}\{{}$\1\6
${}\\{primary}\K\T{0};{}$\6
\&{if} ${}(\\{cur\_col}\E\\{col\_array}+\T{1}){}$\1\5
\\{panic}(\.{"No\ primary\ columns"});\2\6
${}(\\{cur\_col}-\T{1})\MG\\{next}\K{\AND}\\{root},\39\\{root}.\\{prev}\K\\{cur%
\_col}-\T{1};{}$\6
\&{continue};\6
\4${}\}{}$\2\6
\&{for} ${}(\|q\K\|p+\T{1};{}$ ${}\R\\{isspace}({*}\|q);{}$ ${}\|q\PP){}$\1\5
;\2\6
\&{if} ${}(\|q>\|p+\T{7}){}$\1\5
\\{panic}(\.{"Column\ name\ too\ lon}\)\.{g"});\2\6
\&{if} ${}(\\{cur\_col}\G{\AND}\\{col\_array}[\\{max\_cols}]){}$\1\5
\\{panic}(\.{"Too\ many\ columns"});\2\6
\&{for} ${}(\|q\K\\{cur\_col}\MG\\{name};{}$ ${}\R\\{isspace}({*}\|p);{}$ ${}%
\|q\PP,\39\|p\PP){}$\1\5
${}{*}\|q\K{*}\|p;{}$\2\6
${}\\{cur\_col}\MG\\{head}.\\{up}\K\\{cur\_col}\MG\\{head}.\\{down}\K{\AND}%
\\{cur\_col}\MG\\{head};{}$\6
${}\\{cur\_col}\MG\\{len}\K\T{0};{}$\6
\&{if} (\\{primary})\1\5
${}\\{cur\_col}\MG\\{prev}\K\\{cur\_col}-\T{1},\39(\\{cur\_col}-\T{1})\MG%
\\{next}\K\\{cur\_col};{}$\2\6
\&{else}\1\5
${}\\{cur\_col}\MG\\{prev}\K\\{cur\_col}\MG\\{next}\K\\{cur\_col};{}$\2\6
${}\\{cur\_col}\PP;{}$\6
\4${}\}{}$\2\6
\&{if} (\\{primary})\5
${}\{{}$\1\6
\&{if} ${}(\\{cur\_col}\E\\{col\_array}+\T{1}){}$\1\5
\\{panic}(\.{"No\ primary\ columns"});\2\6
${}(\\{cur\_col}-\T{1})\MG\\{next}\K{\AND}\\{root},\39\\{root}.\\{prev}\K\\{cur%
\_col}-\T{1};{}$\6
\4${}\}{}$\2\par
\U7.\fi

\M[203 xgdance.w]{10}\B\X10:Local variables\X${}\E{}$\6
\&{register} \&{column} ${}{*}\\{cur\_col};{}$\6
\&{register} \&{char} ${}{*}\|p,{}$ ${}{*}\|q;{}$\6
\&{register} \&{node} ${}{*}\\{cur\_node};{}$\6
\&{int} \\{primary};\par
\As13\ET20.
\U1.\fi

\M[209 xgdance.w]{11}\B\X11:Read the rows\X${}\E{}$\6
$\\{cur\_node}\K\\{node\_array};{}$\6
\&{while} ${}(\\{fgets}(\\{buf},\39\\{buf\_size},\39\\{stdin})){}$\5
${}\{{}$\1\6
\&{register} \&{column} ${}{*}\\{ccol};{}$\6
\&{register} \&{node} ${}{*}\\{row\_start},{}$ ${}{*}\|x;{}$\7
\&{if} ${}(\\{buf}[\\{strlen}(\\{buf})-\T{1}]\I\.{'\\n'}){}$\1\5
\\{panic}(\.{"Input\ line\ too\ long}\)\.{"});\2\6
${}\\{row\_start}\K\NULL;{}$\6
\&{for} ${}(\|p\K\\{buf};{}$ ${}{*}\|p;{}$ ${}\|p\PP){}$\5
${}\{{}$\1\6
\&{while} ${}(\\{isspace}({*}\|p)){}$\1\5
${}\|p\PP;{}$\2\6
\&{if} ${}(\R{*}\|p){}$\1\5
\&{break};\2\6
\&{for} ${}(\|q\K\|p+\T{1};{}$ ${}\R\\{isspace}({*}\|q)\W{*}\|q\I\.{':'};{}$
${}\|q\PP){}$\1\5
;\2\6
\&{if} ${}(\|q>\|p+\T{7}){}$\1\5
\\{panic}(\.{"Column\ name\ too\ lon}\)\.{g"});\2\6
\&{for} ${}(\|q\K\\{cur\_col}\MG\\{name};{}$ ${}\R\\{isspace}({*}\|p)\W{*}\|p\I%
\.{':'};{}$ ${}\|q\PP,\39\|p\PP){}$\1\5
${}{*}\|q\K{*}\|p;{}$\2\6
${}{*}\|q\K\.{'\\0'};{}$\6
\&{for} ${}(\\{ccol}\K\\{col\_array};{}$ ${}\\{strcmp}(\\{ccol}\MG\\{name},\39%
\\{cur\_col}\MG\\{name});{}$ ${}\\{ccol}\PP){}$\1\5
;\2\6
\&{if} ${}(\\{ccol}\E\\{cur\_col}){}$\1\5
\\{panic}(\.{"Unknown\ column\ name}\)\.{"});\2\6
\&{if} ${}(\\{cur\_node}\E{\AND}\\{node\_array}[\\{max\_nodes}]){}$\1\5
\\{panic}(\.{"Too\ many\ nodes"});\2\6
\&{if} ${}(\R\\{row\_start}){}$\1\5
${}\\{row\_start}\K\\{cur\_node};{}$\2\6
\&{else}\1\5
${}\\{cur\_node}\MG\\{left}\K\\{cur\_node}-\T{1},\39(\\{cur\_node}-\T{1})\MG%
\\{right}\K\\{cur\_node};{}$\2\6
\&{for} ${}(\|x\K\\{row\_start};{}$ ${}\|x\I\\{cur\_node};{}$ ${}\|x\PP){}$\1\6
\&{if} ${}(\|x\MG\\{col}\E\\{ccol}){}$\1\5
\\{panic}(\.{"A\ row\ can't\ use\ a\ c}\)\.{olumn\ twice"});\2\2\6
${}\\{cur\_node}\MG\\{col}\K\\{ccol};{}$\6
${}\\{cur\_node}\MG\\{up}\K\\{ccol}\MG\\{head}.\\{up},\39\\{ccol}\MG\\{head}.%
\\{up}\MG\\{down}\K\\{cur\_node};{}$\6
${}\\{ccol}\MG\\{head}.\\{up}\K\\{cur\_node},\39\\{cur\_node}\MG\\{down}\K{%
\AND}\\{ccol}\MG\\{head};{}$\6
${}\\{ccol}\MG\\{len}\PP;{}$\6
\&{if} ${}({*}\|p\E\.{':'}){}$\1\5
\X24:Read a color restriction\X;\2\6
${}\\{cur\_node}\PP;{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{row\_start}){}$\1\5
\\{panic}(\.{"Empty\ row"});\2\6
${}\\{row\_start}\MG\\{left}\K\\{cur\_node}-\T{1},\39(\\{cur\_node}-\T{1})\MG%
\\{right}\K\\{row\_start};{}$\6
\4${}\}{}$\2\par
\U7.\fi

\N[241 xgdance.w]{1}{12}Backtracking.
Our strategy for generating all exact covers will be to repeatedly
choose always the column that appears to be hardest to cover, namely the
column with shortest list, from all columns that still need to be covered.
And we explore all possibilities via depth-first search.

The neat part of this algorithm is the way the lists are maintained.
Depth-first search means last-in-first-out maintenance of data structures;
and it turns out that we need no auxiliary tables to undelete elements from
lists when backing up. The nodes removed from doubly linked lists remember
their former neighbors, because we do no garbage collection.

The basic operation is ``covering a column.'' This means removing it
from the list of columns needing to be covered, and ``blocking'' its
rows: removing nodes from other lists whenever they belong to a row of
a node in this column's list.

\Y\B\4\X12:Backtrack through all solutions\X${}\E{}$\6
$\\{level}\K\T{0};{}$\6
${}\\{cthresh}\K\.{'a'};{}$\6
\4\\{forward}:\5
\X19:Set \PB{\\{best\_col}} to the best column for branching\X;\6
\\{cover}(\\{best\_col});\6
${}\\{cur\_node}\K\\{choice}[\\{level}]\K\\{best\_col}\MG\\{head}.\\{down};{}$\6
\4\\{advance}:\6
\&{if} ${}(\\{cur\_node}\E{\AND}(\\{best\_col}\MG\\{head})){}$\1\5
\&{goto} \\{backup};\2\6
\&{if} ${}(\\{verbose}>\T{1}){}$\5
${}\{{}$\1\6
${}\\{printf}(\.{"L\%d:"},\39\\{level});{}$\6
\\{print\_row}(\\{cur\_node});\6
\4${}\}{}$\2\6
${}\\{conflict}\K\T{0};{}$\6
\X17:Cover all other columns of \PB{\\{cur\_node}}\X;\6
\&{if} (\\{conflict})\1\5
\&{goto} \\{recover};\2\6
\&{if} ${}(\\{root}.\\{next}\E{\AND}\\{root}){}$\1\5
\X21:Record solution and \PB{\&{goto} \\{recover}}\X;\2\6
${}\\{level}\PP;{}$\6
\&{goto} \\{forward};\6
\4\\{backup}:\5
\\{uncover}(\\{best\_col});\6
\&{if} ${}(\\{level}\E\T{0}){}$\1\5
\&{goto} \\{done};\2\6
${}\\{level}\MM;{}$\6
${}\\{cur\_node}\K\\{choice}[\\{level}]{}$;\5
${}\\{best\_col}\K\\{cur\_node}\MG\\{col};{}$\6
\4\\{recover}:\5
\X18:Uncover all other columns of \PB{\\{cur\_node}}\X;\6
${}\\{cur\_node}\K\\{choice}[\\{level}]\K\\{cur\_node}\MG\\{down}{}$;\5
\&{goto} \\{advance};\6
\4\\{done}:\6
\&{if} ${}(\\{verbose}>\T{3}){}$\1\5
\X22:Print column lengths, to make sure everything has been restored\X;\2\par
\U1.\fi

\M[284 xgdance.w]{13}\B\X10:Local variables\X${}\mathrel+\E{}$\6
\&{register} \&{column} ${}{*}\\{best\_col}{}$;\C{ column chosen for branching
}\6
\&{register} \&{node} ${}{*}\\{pp}{}$;\C{ traverses a row }\par
\fi

\M[288 xgdance.w]{14}\B\X2:Global variables\X${}\mathrel+\E{}$\6
\&{int} \\{level};\C{ number of choices in current partial solution }\6
\&{int} \\{cthresh};\C{ smallest color allowable when extending a solution }\6
\&{int} \\{conflict};\C{ set nonzero if a conflict arises while covering }\6
\&{node} ${}{*}\\{choice}[\\{max\_level}]{}$;\C{ the row and column chosen on
each level }\par
\fi

\M[294 xgdance.w]{15}When a row is blocked, it leaves all lists except the list
of the
column that is being covered. Thus a node is never removed from a list
twice.

\Y\B\4\X6:Subroutines\X${}\mathrel+\E{}$\6
\\{cover}(\|c)\1\1\6
\&{column} ${}{*}\|c;\2\2{}$\6
${}\{{}$\5
\1\&{register} \&{column} ${}{*}\|l,{}$ ${}{*}\|r;{}$\6
\&{register} \&{node} ${}{*}\\{rr},{}$ ${}{*}\\{nn},{}$ ${}{*}\\{uu},{}$ ${}{*}%
\\{dd};{}$\6
\&{register} \&{int} \|k${}\K\T{1}{}$;\C{ updates }\7
${}\|l\K\|c\MG\\{prev}{}$;\5
${}\|r\K\|c\MG\\{next};{}$\6
${}\|l\MG\\{next}\K\|r{}$;\5
${}\|r\MG\\{prev}\K\|l;{}$\6
\&{for} ${}(\\{rr}\K\|c\MG\\{head}.\\{down};{}$ ${}\\{rr}\I{\AND}(\|c\MG%
\\{head});{}$ ${}\\{rr}\K\\{rr}\MG\\{down}){}$\1\6
\&{for} ${}(\\{nn}\K\\{rr}\MG\\{right};{}$ ${}\\{nn}\I\\{rr};{}$ ${}\\{nn}\K%
\\{nn}\MG\\{right}){}$\5
${}\{{}$\1\6
${}\\{uu}\K\\{nn}\MG\\{up}{}$;\5
${}\\{dd}\K\\{nn}\MG\\{down};{}$\6
${}\\{uu}\MG\\{down}\K\\{dd}{}$;\5
${}\\{dd}\MG\\{up}\K\\{uu};{}$\6
${}\|k\PP;{}$\6
${}\\{nn}\MG\\{col}\MG\\{len}\MM;{}$\6
\4${}\}{}$\2\2\6
${}\\{updates}\MRL{+{\K}}\|k;{}$\6
${}\\{upd\_prof}[\\{level}]\MRL{+{\K}}\|k;{}$\6
\4${}\}{}$\2\par
\fi

\M[317 xgdance.w]{16}Uncovering is done in precisely the reverse order. The
pointers thereby
execute an exquisitely choreo\-graphed dance which returns them almost
magically to their former state.

\Y\B\4\X6:Subroutines\X${}\mathrel+\E{}$\6
\\{uncover}(\|c)\1\1\6
\&{column} ${}{*}\|c;\2\2{}$\6
${}\{{}$\5
\1\&{register} \&{column} ${}{*}\|l,{}$ ${}{*}\|r;{}$\6
\&{register} \&{node} ${}{*}\\{rr},{}$ ${}{*}\\{nn},{}$ ${}{*}\\{uu},{}$ ${}{*}%
\\{dd};{}$\7
\&{for} ${}(\\{rr}\K\|c\MG\\{head}.\\{up};{}$ ${}\\{rr}\I{\AND}(\|c\MG%
\\{head});{}$ ${}\\{rr}\K\\{rr}\MG\\{up}){}$\1\6
\&{for} ${}(\\{nn}\K\\{rr}\MG\\{left};{}$ ${}\\{nn}\I\\{rr};{}$ ${}\\{nn}\K%
\\{nn}\MG\\{left}){}$\5
${}\{{}$\1\6
${}\\{uu}\K\\{nn}\MG\\{up}{}$;\5
${}\\{dd}\K\\{nn}\MG\\{down};{}$\6
${}\\{uu}\MG\\{down}\K\\{dd}\MG\\{up}\K\\{nn};{}$\6
${}\\{nn}\MG\\{col}\MG\\{len}\PP;{}$\6
\4${}\}{}$\2\2\6
${}\|l\K\|c\MG\\{prev}{}$;\5
${}\|r\K\|c\MG\\{next};{}$\6
${}\|l\MG\\{next}\K\|r\MG\\{prev}\K\|c;{}$\6
\4${}\}{}$\2\par
\fi

\M[336 xgdance.w]{17}\B\X17:Cover all other columns of \PB{\\{cur\_node}}\X${}%
\E{}$\6
\&{for} ${}(\\{pp}\K\\{cur\_node}\MG\\{right};{}$ ${}\\{pp}\I\\{cur\_node};{}$
${}\\{pp}\K\\{pp}\MG\\{right}){}$\1\6
\&{if} ${}(\R\\{pp}\MG\\{color}){}$\1\5
${}\\{cover}(\\{pp}\MG\\{col});{}$\2\6
\&{else} \&{if} ${}(\\{pp}\MG\\{color}>\T{0}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{pp}\MG\\{color}>\\{cthresh}){}$\1\5
${}\\{conflict}\K\T{1};{}$\2\6
\&{else}\1\5
\\{purify}(\\{pp});\2\6
\4${}\}{}$\2\2\par
\U12.\fi

\M[344 xgdance.w]{18}We included \PB{\\{left}} links, thereby making the rows
doubly linked, so
that columns would be uncovered in the correct LIFO order in this
part of the program. (The \PB{\\{uncover}} routine itself could have done its
job with \PB{\\{right}} links only.) (Think about it.)

\Y\B\4\X18:Uncover all other columns of \PB{\\{cur\_node}}\X${}\E{}$\6
\&{for} ${}(\\{pp}\K\\{cur\_node}\MG\\{left};{}$ ${}\\{pp}\I\\{cur\_node};{}$
${}\\{pp}\K\\{pp}\MG\\{left}){}$\1\6
\&{if} ${}(\R\\{pp}\MG\\{color}){}$\1\5
${}\\{uncover}(\\{pp}\MG\\{col});{}$\2\6
\&{else} \&{if} ${}(\\{pp}\MG\\{color}>\T{0}\W\\{pp}\MG\\{color}\Z%
\\{cthresh}){}$\1\5
\\{unpurify}(\\{pp});\2\2\par
\U12.\fi

\M[354 xgdance.w]{19}\B\X19:Set \PB{\\{best\_col}} to the best column for
branching\X${}\E{}$\6
$\\{minlen}\K\\{max\_nodes};{}$\6
\&{if} ${}(\\{verbose}>\T{2}){}$\1\5
${}\\{printf}(\.{"Level\ \%d:"},\39\\{level});{}$\2\6
\&{for} ${}(\\{cur\_col}\K\\{root}.\\{next};{}$ ${}\\{cur\_col}\I{\AND}%
\\{root};{}$ ${}\\{cur\_col}\K\\{cur\_col}\MG\\{next}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{verbose}>\T{2}){}$\1\5
${}\\{printf}(\.{"\ \%s(\%d)"},\39\\{cur\_col}\MG\\{name},\39\\{cur\_col}\MG%
\\{len});{}$\2\6
\&{if} ${}(\\{cur\_col}\MG\\{len}<\\{minlen}){}$\1\5
${}\\{best\_col}\K\\{cur\_col},\39\\{minlen}\K\\{cur\_col}\MG\\{len};{}$\2\6
\4${}\}{}$\2\6
\&{if} (\\{verbose})\5
${}\{{}$\1\6
\&{if} ${}(\\{level}>\\{maxl}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{level}\G\\{max\_level}){}$\1\5
\\{panic}(\.{"Too\ many\ levels"});\2\6
${}\\{maxl}\K\\{level};{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{minlen}>\\{maxb}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{minlen}\G\\{max\_degree}){}$\1\5
\\{panic}(\.{"Too\ many\ branches"});\2\6
${}\\{maxb}\K\\{minlen};{}$\6
\4${}\}{}$\2\6
${}\\{profile}[\\{level}][\\{minlen}]\PP;{}$\6
\&{if} ${}(\\{verbose}>\T{2}){}$\1\5
${}\\{printf}(\.{"\ branching\ on\ \%s(\%d}\)\.{)\\n"},\39\\{best\_col}\MG%
\\{name},\39\\{minlen});{}$\2\6
\4${}\}{}$\2\par
\U12.\fi

\M[374 xgdance.w]{20}\B\X10:Local variables\X${}\mathrel+\E{}$\6
\&{register} \&{int} \\{minlen};\6
\&{register} \&{int} \|j${},{}$ \|k${},{}$ \|x;\par
\fi

\M[378 xgdance.w]{21}\B\X21:Record solution and \PB{\&{goto} \\{recover}}\X${}%
\E{}$\6
${}\{{}$\1\6
${}\\{count}\PP;{}$\6
\&{if} (\\{verbose})\5
${}\{{}$\1\6
${}\\{profile}[\\{level}+\T{1}][\T{0}]\PP;{}$\6
\&{if} ${}(\\{count}\MOD\\{spacing}\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{printf}(\.{"\%d:\\n"},\39\\{count});{}$\6
\&{for} ${}(\|k\K\T{0};{}$ ${}\|k\Z\\{level};{}$ ${}\|k\PP){}$\1\5
\\{print\_row}(\\{choice}[\|k]);\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{goto} \\{recover};\6
\4${}\}{}$\2\par
\U12.\fi

\M[391 xgdance.w]{22}\B\X22:Print column lengths, to make sure everything has
been restored\X${}\E{}$\6
${}\{{}$\1\6
\\{printf}(\.{"Final\ column\ length}\)\.{s"});\6
\&{for} ${}(\\{cur\_col}\K\\{root}.\\{next};{}$ ${}\\{cur\_col}\I{\AND}%
\\{root};{}$ ${}\\{cur\_col}\K\\{cur\_col}\MG\\{next}){}$\1\5
${}\\{printf}(\.{"\ \%s(\%d)"},\39\\{cur\_col}\MG\\{name},\39\\{cur\_col}\MG%
\\{len});{}$\2\6
\\{printf}(\.{"\\n"});\6
\4${}\}{}$\2\par
\U12.\fi

\M[399 xgdance.w]{23}\B\X23:Print a profile of the search tree\X${}\E{}$\6
${}\{{}$\1\6
${}\|x\K\T{1}{}$;\C{ the root node doesn't show up in the profile }\6
\&{for} ${}(\\{level}\K\T{1};{}$ ${}\\{level}\Z\\{maxl}+\T{1};{}$ ${}\\{level}%
\PP){}$\5
${}\{{}$\1\6
${}\|j\K\T{0};{}$\6
\&{for} ${}(\|k\K\T{0};{}$ ${}\|k\Z\\{maxb};{}$ ${}\|k\PP){}$\5
${}\{{}$\1\6
${}\\{printf}(\.{"\%6d"},\39\\{profile}[\\{level}][\|k]);{}$\6
${}\|j\MRL{+{\K}}\\{profile}[\\{level}][\|k];{}$\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"\%8d\ nodes,\ \%u\ updat}\)\.{es,\ \%u\ cleansings\\n"},\39%
\|j,\39\\{upd\_prof}[\\{level}-\T{1}],\39\\{pur\_prof}[\\{level}-\T{1}]);{}$\6
${}\|x\MRL{+{\K}}\|j;{}$\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"Total\ \%d\ nodes.\\n"},\39\|x);{}$\6
\4${}\}{}$\2\par
\U1.\fi

\N[415 xgdance.w]{1}{24}Color barriers. Finally, here's the new material
related to coloring.

\Y\B\4\X24:Read a color restriction\X${}\E{}$\6
${}\{{}$\1\6
\&{if} (\\{primary})\1\5
\\{panic}(\.{"Color\ isn't\ allowed}\)\.{\ in\ a\ primary\ column}\)\.{"});\2\6
\&{if} ${}(\\{isspace}({*}(\|p+\T{1}))\V\R\\{isspace}({*}(\|p+\T{2}))){}$\1\5
\\{panic}(\.{"Color\ should\ be\ a\ s}\)\.{ingle\ character"});\2\6
${}\\{cur\_node}\MG\\{color}\K{*}(\|p+\T{1});{}$\6
${}\|p\MRL{+{\K}}\T{2};{}$\6
\4${}\}{}$\2\par
\U11.\fi

\M[426 xgdance.w]{25}When we choose a row that specifies colors in one or more
columns, we
``purify'' those columns by removing all incompatible rows. All rows that
want the same color in a purified column will now be given the color code~$-1$
so that we need not purify the column again.

\Y\B\4\X6:Subroutines\X${}\mathrel+\E{}$\6
\\{purify}(\|p)\1\1\6
\&{node} ${}{*}\|p;\2\2{}$\6
${}\{{}$\5
\1\&{register} \&{column} ${}{*}\|c\K\|p\MG\\{col};{}$\6
\&{register} \&{int} \|x${}\K\|p\MG\\{color};{}$\6
\&{register} \&{node} ${}{*}\\{rr},{}$ ${}{*}\\{nn},{}$ ${}{*}\\{uu},{}$ ${}{*}%
\\{dd};{}$\6
\&{register} \&{int} \|k${}\K\T{0},{}$ \\{kk}${}\K\T{1}{}$;\C{ updates }\7
${}\|c\MG\\{head}.\\{color}\K\|x{}$;\C{ this is used only to help \PB{\\{print%
\_row}} }\6
${}\|c\MG\\{color\_thresh}\K\\{cthresh};{}$\6
\&{if} ${}(\\{cthresh}\E\|x){}$\1\5
${}\\{cthresh}\PP;{}$\2\6
\&{for} ${}(\\{rr}\K\|c\MG\\{head}.\\{down};{}$ ${}\\{rr}\I{\AND}(\|c\MG%
\\{head});{}$ ${}\\{rr}\K\\{rr}\MG\\{down}){}$\1\6
\&{if} ${}(\\{rr}\MG\\{color}\I\|x){}$\5
${}\{{}$\1\6
\&{for} ${}(\\{nn}\K\\{rr}\MG\\{right};{}$ ${}\\{nn}\I\\{rr};{}$ ${}\\{nn}\K%
\\{nn}\MG\\{right}){}$\5
${}\{{}$\1\6
${}\\{uu}\K\\{nn}\MG\\{up}{}$;\5
${}\\{dd}\K\\{nn}\MG\\{down};{}$\6
${}\\{uu}\MG\\{down}\K\\{dd}{}$;\5
${}\\{dd}\MG\\{up}\K\\{uu};{}$\6
${}\|k\PP;{}$\6
${}\\{nn}\MG\\{col}\MG\\{len}\MM;{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\5
\2\&{else} \&{if} ${}(\\{rr}\I\|p){}$\1\5
${}\\{kk}\PP,\39\\{rr}\MG\\{color}\K{-}\T{1};{}$\2\2\6
${}\\{updates}\MRL{+{\K}}\|k,\39\\{purifs}\MRL{+{\K}}\\{kk};{}$\6
${}\\{upd\_prof}[\\{level}]\MRL{+{\K}}\|k,\39\\{pur\_prof}[\\{level}]\MRL{+{%
\K}}\\{kk};{}$\6
\4${}\}{}$\2\par
\fi

\M[454 xgdance.w]{26}Just as \PB{\\{purify}} is analogous to \PB{\\{cover}},
the inverse process is analogous
to \PB{\\{uncover}}.

\Y\B\4\X6:Subroutines\X${}\mathrel+\E{}$\6
\\{unpurify}(\|p)\1\1\6
\&{node} ${}{*}\|p;\2\2{}$\6
${}\{{}$\5
\1\&{register} \&{column} ${}{*}\|c\K\|p\MG\\{col};{}$\6
\&{register} \&{int} \|x${}\K\|p\MG\\{color};{}$\6
\&{register} \&{node} ${}{*}\\{rr},{}$ ${}{*}\\{nn},{}$ ${}{*}\\{uu},{}$ ${}{*}%
\\{dd};{}$\7
\&{for} ${}(\\{rr}\K\|c\MG\\{head}.\\{up};{}$ ${}\\{rr}\I{\AND}(\|c\MG%
\\{head});{}$ ${}\\{rr}\K\\{rr}\MG\\{up}){}$\1\6
\&{if} ${}(\\{rr}\MG\\{color}<\T{0}){}$\1\5
${}\\{rr}\MG\\{color}\K\|x;{}$\2\6
\&{else} \&{if} ${}(\\{rr}\I\|p){}$\5
${}\{{}$\1\6
\&{for} ${}(\\{nn}\K\\{rr}\MG\\{left};{}$ ${}\\{nn}\I\\{rr};{}$ ${}\\{nn}\K%
\\{nn}\MG\\{left}){}$\5
${}\{{}$\1\6
${}\\{uu}\K\\{nn}\MG\\{up}{}$;\5
${}\\{dd}\K\\{nn}\MG\\{down};{}$\6
${}\\{uu}\MG\\{down}\K\\{dd}\MG\\{up}\K\\{nn};{}$\6
${}\\{nn}\MG\\{col}\MG\\{len}\PP;{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\2\6
${}\|c\MG\\{head}.\\{color}\K\T{0};{}$\6
${}\\{cthresh}\K\|c\MG\\{color\_thresh};{}$\6
\4${}\}{}$\2\par
\fi

\M[476 xgdance.w]{27}\B\X27:Print the color of node \PB{\|q}\X${}\E{}$\6
$\\{printf}(\.{":\%c"},\39\|q\MG\\{color}>\T{0}\?\|q\MG\\{color}:\|q\MG\\{col}%
\MG\\{head}.\\{color}){}$;\par
\U6.\fi

\N[479 xgdance.w]{1}{28}Help for debugging. Here's a subroutine for when I'm
doing a long run and want to check the current progress.

\Y\B\4\X6:Subroutines\X${}\mathrel+\E{}$\6
\&{void} \\{show\_state}(\,)\1\1\2\2\6
${}\{{}$\1\6
\&{register} \&{int} \|k;\7
${}\\{printf}(\.{"Current\ state\ (leve}\)\.{l\ \%d):\\n"},\39\\{level});{}$\6
\&{for} ${}(\|k\K\T{0};{}$ ${}\|k<\\{level};{}$ ${}\|k\PP){}$\1\5
\\{print\_row}(\\{choice}[\|k]);\2\6
${}\\{printf}(\.{"Max\ level\ so\ far:\ \%}\)\.{d\\n"},\39\\{maxl});{}$\6
${}\\{printf}(\.{"Max\ branching\ so\ fa}\)\.{r:\ \%d\\n"},\39\\{maxb});{}$\6
${}\\{printf}(\.{"Solutions\ so\ far:\ \%}\)\.{d\\n"},\39\\{count});{}$\6
\4${}\}{}$\2\par
\fi

\N[493 xgdance.w]{1}{29}Index.
\fi

\inx
\fin
\con
